%% English translation of effect-systems-free-variables.tex.
%% Standalone article version; build with xelatex + biber:
%%   latexmk -pdf -xelatex effect-systems-free-variables-en.tex
%% (xelatex is used so the < > | glyphs in the code listings render literally,
%%  exactly as in the Russian original; bib.bib is reused verbatim.)
\documentclass[11pt]{article}

\usepackage[a4paper, left=2.4cm, right=2.4cm, top=2.4cm, bottom=2.6cm]{geometry}
\usepackage{fontspec}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{booktabs}
\usepackage{array}
\usepackage{listings}
\usepackage{tikz}
\usetikzlibrary{positioning, arrows.meta}
\usepackage[english]{babel}
\usepackage{csquotes}
\usepackage{hyperref}
\usepackage[backend=biber, style=numeric, sorting=none, maxbibnames=99]{biblatex}
\addbibresource{bib.bib}

% --- listings: same lightweight language definitions as the original ---
\lstdefinestyle{modern}{
    basicstyle=\ttfamily\footnotesize,
    keywordstyle=\bfseries,
    commentstyle=\itshape,
    stringstyle=,
    numberstyle=\tiny,
    tabsize=2,
    showspaces=false,
    showstringspaces=false,
    breaklines=true,
    breakatwhitespace=true,
    captionpos=b,
    escapeinside={(*@}{@*)},
    columns=fullflexible,
    keepspaces=true,
}
\lstset{style=modern}
\lstdefinelanguage{colang}{
    keywords={let,var,effect,perform,data,context,fun,where,if,else,throw,try,catch,local,free,op,handle,resume,return},
    sensitive=true, comment=[l]{//}, morecomment=[s]{/*}{*/}, morestring=[b]",
}
\lstdefinelanguage{modal}{
    keywords={fun,let,if,else,handle,with,mask},
    sensitive=true, comment=[l]{//}, morestring=[b]",
}

\title{Effect Systems as a Mechanism for Managing\\ Contextual Free Variables: A Unifying Perspective}
\author{A.\,S. Stoyan\thanks{National Research University Higher School of Economics,
3A Kantemirovskaya Str., Saint Petersburg, 194100, Russia.
Email: \texttt{a.stoyan@hse.ru}. ORCID: 0009-0004-6935-1447.}}
\date{}

\begin{document}
\maketitle

\begin{abstract}
Effect systems lift information about a program's computational effects to the type
level: they sharpen abstraction boundaries and let the compiler verify that an effectful
computation runs in a context that supports it. Nevertheless, static effect tracking has
not yet become common practice: the leading approaches --- effect rows, capabilities, and
modal types --- have evolved as unrelated constructions, and a language designer
struggles to see the overall design space. We propose a unifying perspective: an effect
system is a mechanism for managing the contextual free variables of a computation,
i.e., dependencies that are not passed as ordinary arguments but must be resolved by the
execution context. Such a variable goes through a lifecycle of two phases: it arises
pending, as an unfulfilled contextual requirement, and once bound, its value can be
captured by a closure as a lexical free variable. Families of effect systems differ in
which phases and transitions of this lifecycle they make explicit in types. Classical
row-based systems (Koka, Links, Java's checked exceptions) keep the variables pending and
cannot express capture. Capability-based systems allow capture but forbid a captured
variable from escaping its context by means of type-level escape analysis. Modal systems,
instead of forbidding the escape, release the escaped variable again, re-exposing it in
the type as a contextual requirement. Comparing the families along this single axis shows
that the static discipline of an effect system can be assembled from two mechanisms
already familiar to practitioners: implicit parameters and type-level escape analysis.
\end{abstract}

\noindent\textbf{Keywords:} effect systems; computational effects; contextual free
variables; free variables; implicit parameters; escape analysis; capabilities; effect
rows; modal types; type systems; contextual polymorphism.

\bigskip


\section{Introduction}

Programming is, to a large extent, about managing complexity: a developer strives to
focus on an individual fragment of behaviour without holding the whole program in mind at
once. Abstraction and separation of concerns are the key devices: different
responsibilities are delegated to different entities, such as functions or modules. It is
often convenient to single out, as such an entity, the \emph{execution context} of a
computation, while \emph{effects} arise naturally as the interactions of the computation
with that context~\cite{kiselyov2013extensible}.

Let us start from the opposite notion --- purity. The only observable result of a pure
computation is its value. A function is called pure if it returns the same result on the
same arguments, and its application is a pure computation. Programming with pure functions
is generally considered good practice: the composition of pure functions is pure, the
semantics of the code is transparent, and type systems work well, providing partial
specification, safety, and clarity of abstractions.

However, writing and maintaining a non-trivial program using pure computations alone is
hard. All state and control flow must be managed explicitly through the arguments and
results of functions. As an example of manual error propagation: a function returns a
\lstinline|Result|, and every call is accompanied by a check. What is needed is a tool
for managing complexity --- something that hides such inessential details.

\begin{lstlisting}[language=colang]
    fun readInts(n: Int): Result<List<Int>, Error> {
        if (n == 0) { return Ok(Nil()) }
        let x = readInt()
        if (x is Err) { return x }
        let xs = readInts(n - 1)
        if (xs is Err) { return xs }
        Ok(Node(x.value, xs.value))
    }
\end{lstlisting}

Computational effects are exactly such a tool. An effect is an interaction of a
computation with its execution context; the computation addresses the context directly,
without the ceremony that pervades pure code. The context may be the run-time system
providing OS services, mutable memory cells, control-transfer mechanisms, input/output,
and so on. An \emph{effectful operation} is a construct that gives rise to an effect. For
example, the exception mechanism lets a computation report an error directly to the
context --- the nearest handler: the routine propagation of errors disappears from the
code, and the essence remains.

\noindent\begin{minipage}{\linewidth}
\begin{lstlisting}[language=colang]
    fun readInts(n: Int): List<Int> {
        if (n == 0) { return Nil() }
        Node(readInt(), readInts(n - 1))
    }
\end{lstlisting}
\end{minipage}

Some approaches let the programmer define their own effects and execution contexts as
library code. In particular, \emph{effect handlers} are a promising approach that has
proven itself beyond the community of purely functional
languages~\cite{plotkin2013handling, chandrasekaran2018algebraic}. Handlers separate the
signature of an effect from its implementation~\cite{kiselyov2013extensible}: an
expression can be viewed as a client that evaluates either to a pure result or to a
request to a server (the execution context, the effect handler). A handler delimits a
region in which certain effectful operations are possible, and interprets them in terms
of pure values and the effects of the enclosing region. User-defined effects serve as a
tool for building embedded domain-specific languages~\cite{leijen2017structured}, and
various such approaches compete in expressiveness, performance, and the convenience of
composing independently defined effects~\cite{liang1995monad, kiselyov2013extensible,
schrijvers2019monad, van2024framework}.

Many effectful operations are admissible only within a limited region. For example, an
exception can be thrown only within the context of a corresponding handler, and a mutable
cell allocated on the stack cannot be used after the function returns. Static control over
the use of such operations prevents many run-time errors and improves program
reliability. \emph{Effect systems} are precisely the mechanism for providing such control
at the type level. Moreover, since effects are inherently implicit at the term level, it
is hard to judge their presence from the source code: thus the signature of the effectful
version of \lstinline|readInts| above is silent about the possible error. Some degree of
explicitness is therefore still required, which effect systems provide at the type level.
This helps to state clear functional abstractions and to prevent implementation details
from silently leaking as interactions with the context.

The idea of tracking effects in types goes back to the work of Lucassen and
Gifford~\cite{lucassen1988polymorphic}; the counter-movement of the modern mainstream is
also telling: OCaml~5 has adopted effect handlers without tracking them in types at
all~\cite{sivaramakrishnan2021retrofitting}, which only sharpens the question of a
practical effect system. Many promising effect-system constructions have been proposed:
row-polymorphic types~\cite{leijen2014koka}, capabilities~\cite{brachthauser2022effects,
boruch2023capturing}, modal types~\cite{tang2025modal}, and others. However, each approach
has its own strengths and weaknesses, and the overall picture of the design space is
insufficiently studied. Yet a language designer needs a holistic view of the available
options in order to choose the most suitable solution given the features and constraints
of the target language.

In this work we propose a unifying view of these seemingly disparate approaches:
\emph{an effect system is a mechanism for managing the contextual free variables of a
computation at the type level}. Such variables are not passed as ordinary term arguments
but must be resolved by the surrounding context: a handler, a resource, a capability, or
another witness. This view, which draws on observations from the Scala
language~\cite{odersky2021safer, boruch2023capturing}, lets us regard the classical type
system and the effect system as two sides of one problem: the former manages
\emph{bound} variables, the latter --- contextual free variables. These variables, in
turn, pass through a single \emph{lifecycle}, and each existing family of effect systems
turns out to be a choice of which phases and transitions of this cycle are available and
how they are reflected at the type level.

This work is precisely a unifying perspective, not an exhaustive survey: the scope is
deliberately narrowed to three families, selected by a practical criterion ---
constructions embodied in programming languages and aimed at the application programmer.
Most literally, this perspective applies to effects arising from interaction with the
execution context: algebraic operations, exceptions, access to resources and
capabilities. For effect predicates on computations --- for example, divergence, cost
estimates, or information-security labels --- the reading through contextual free
variables is more of an analogy than a literal model.

This work is based on our earlier English-language survey~\cite{stoyan2024modern} but
differs from it in substance. New are the lifecycle model of a contextual free variable
itself (Fig.~\ref{fig:lifecycle}, Table~\ref{tbl:free-vars}), the notion of
\emph{uncapturing} of a captured variable, and the reading of each family in terms of the
cycle (Sections~\ref{sec:rows}--\ref{sec:modal}).

The contributions of the work can be summarized as follows:
\begin{itemize}
    \item we formulate a single point of view on effect systems as a means of managing
    contextual free variables and describe their \emph{lifecycle}, along whose phases and
    transitions all the constructions under consideration are arranged
    (Section~\ref{sec:idea});
    \item we revisit three families of modern effect systems --- row-based
    (Section~\ref{sec:rows}), capability-based (Section~\ref{sec:capabilities}), and modal
    (Section~\ref{sec:modal}) --- and show which transitions of the cycle each of them
    supports;
    \item we carry out a comparative analysis of the families along a single lifecycle
    axis and show that the static discipline of such an effect system can be regarded as a
    combination of two mechanisms already familiar to practitioners --- implicit
    parameters and type-level escape analysis (Sections~\ref{sec:comparison}
    and~\ref{subsec:synthesis}).
\end{itemize}


\section{Free Variables as an Embodiment of Requests to the Context}
\label{sec:es-intro}

In this section we show that the pragmatic motives of effect systems arise naturally from
ordinary programming scenarios, and we prepare the ground for the central idea of the work
(Section~\ref{sec:idea}). Rather than starting from the design of existing systems, we
will derive an effect system constructively --- by a successive refactoring of an everyday
scenario. For informal explanations we will use examples in a fictional language in the
spirit of Scala and Kotlin, introducing its constructs as needed.

Consider a running example: a business-logic function performs a query to a database. In
the first version it establishes the connection itself:
\begin{lstlisting}[language=colang]
    fun business1() {
        let db = Db.open(...)
        db.query("...")
    }
\end{lstlisting}
This solution has drawbacks. First, testability is limited: substituting a stub for the
connection is not easy. Second, the function produces observable side effects ---
accesses to the database --- which are part of its external contract but are in no way
reflected in the signature.

The second version is noticeably better: the function is parameterized by the connection,
and the call site can pass a concrete implementation. Thus the function's behaviour begins
to be determined by the execution context.
\begin{lstlisting}[language=colang]
    fun business2(db: Db) {
        db.query("...")
    }
    fun caller() {
        let db = Db.open(...)
        business2(db)
    }
\end{lstlisting}
Now the observable effects are explicitly expressed through the parameter \lstinline|db|.
However, this comes at the price of syntactic noise at the call site: the administrative
effort of explicitly passing parameters grows. For a single argument this is a trifle, but
if four such parameters must be threaded through several levels of the call stack, the
burden becomes tangible. A practice that aspires to wide adoption must be no less
convenient than lower-quality alternatives.

To cut down the routine, in the third version the ordinary parameter is replaced by a
\emph{dynamic free variable} \lstinline|?db|, which obtains its value by lookup in the
dynamic scope of the call.
\begin{lstlisting}[language=colang]
    fun business3() {
        ?db.query("...")
    }
    fun caller() {
        let ?db = Db.open(...)
        business3()
    }
\end{lstlisting}
Thus, through the free variable \lstinline|?db| the function \lstinline|business3|
requests functionality from the call context.

Nevertheless, the absence of syntactic noise is achieved at the cost of losing other
advantages: the static check of the presence and type of such a variable and its explicit
representation in the signature are lost. This is where effect systems come into play:
they offer ways to preserve static safety without introducing extra ceremony at the term
level. We consider concrete solutions in what follows.

Let us fix the main requirement on an effect system --- \emph{effect safety}: every effect
may be performed only in a context that supports it, otherwise the program must be
rejected. In particular, every exception must be caught, and in general every effect ---
handled. In terms of free variables, this means that every such variable must be resolved
by the call context of the function.

It is also useful to control what happens to such a variable after it has received a
value. A closure may fix that value already as a \emph{lexically bound} free variable and
outlive the context that supplied the value:
\begin{lstlisting}[language=colang]
    fun leak(): () => Unit {
        let f =
            () => ?db.query("...") // the closure fixes the value of ?db
        return f // f outlives the context that gave the value
    }
\end{lstlisting}
Since, as discussed, many effects are admissible only within a limited region, such
captured variables must not leak beyond their contexts --- otherwise effect safety would
be lost; to this end an effect system must support some form of escape
analysis~\cite{park1992escape}. In return, tracking lexical capture makes it possible not
to propagate information about the free variables of an argument function into the type of
a higher-order function --- this technique is called \emph{contextual polymorphism} and
will be considered later (Section~\ref{sec:capabilities}).


\section{The Lifecycle of a Contextual Free Variable}
\label{sec:idea}

The preceding discussion has essentially been about free variables of a special kind. We
will call them \emph{contextual free variables} and understand them broadly: these are
dependencies of a computation on the surrounding execution context --- an effect handler,
a capability value, a resource, or another witness, without which the computation cannot
be safely run. The word \enquote{variable} here fixes not a mandatory syntactic
representation but a role in the program: just as a free variable of a term must be
resolved by its environment, so an effectful computation requires support from the
context. We will call the obligation of the context to satisfy such a dependency a
\emph{contextual requirement}; it is the type-level trace of a not-yet-resolved variable
--- a technical name for what the type system must track. In different systems the same
role is represented differently --- as an operation label in a row, as an implicit
parameter, as a capability value, or as a modal requirement.

Our view grows out of a simple observation: such a variable may pass through a
\emph{lifecycle} of two phases --- \emph{pending} and \emph{capture}
(Fig.~\ref{fig:lifecycle}). The variable arises \emph{pending}: the body of the function
needs something from the context but has not yet received it. Then the pending state is
\emph{fulfilled} --- the calling context provides a suitable value, as for the variable
\lstinline|?db| from Section~\ref{sec:es-intro} --- and, once it enters a deferred
computation, this value becomes \emph{captured} by a closure; the captured variable must
be \emph{retained} within the region of its context, which is checked by type-level escape
analysis. Here and throughout, by escape analysis we mean the static control over whether
a value that has captured such a variable may outlive the context that bound it. Finally,
upon crossing the boundary of the context, the captured variable may be \emph{released
again} --- moved back into the pending phase and once more reflected in the type as a
contextual requirement (this is what modal systems do, Section~\ref{sec:modal}). Note that
\enquote{fulfilled} is not a separate phase but a transition: a bound but not captured
variable lives only inside a running activation of the function and does not enter values,
so the types need not track it.

The two ends of this cycle correspond to two classical disciplines of resolving free
variables. The pending phase is dynamic: the value of the variable (for example, a
handler) is determined by the calling context. The capture phase is lexical: the value has
already been obtained and fixed at the point of capture by the closure. For brevity we
will call a variable in the pending phase \emph{dynamic}, and a captured one
\emph{lexical}. The correspondence of the two phases is summarized in
Table~\ref{tbl:free-vars}.

Let us note a connection with a known notion: a pending contextual requirement is, in
essence, a \emph{coeffect} in the sense of works on the context-dependence of
computations~\cite{petricek2014coeffects}, whose canonical examples are implicit
parameters and dynamic binding. The orthogonal contribution of the present work is the
\emph{capture} layer, with escape analysis and uncapturing, which standard coeffect theory
does not describe.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  node distance=1.0cm and 3.2cm,
  phase/.style={draw, thick, rounded corners, align=center, font=\footnotesize,
                inner sep=4pt, minimum width=2.4cm, minimum height=0.95cm},
  lbl/.style={font=\footnotesize, align=center}]
  \node[phase] (unf) {Pending};
  \node[phase, right=of unf] (cap) {Captured};
  \draw[thick, -{Latex}] (unf) to[bend left=18] node[lbl, above] {fulfilment and capture by a closure} (cap);
  \draw[thick, -{Latex}] (cap) to[loop right] node[lbl, right] {retention\\(escape analysis)} (cap);
  \draw[thick, -{Latex}] (cap) to[bend left=18] node[lbl, below] {uncapturing} (unf);
\end{tikzpicture}
\caption{Lifecycle of a contextual free variable}
\label{fig:lifecycle}
\end{figure}

\begin{table}[t]
  \caption{Two phases of a contextual free variable's lifecycle}
  \label{tbl:free-vars}
  \centering
  \begin{tabular}{>{\raggedright\arraybackslash}p{3.3cm}>{\raggedright\arraybackslash}p{4.6cm}>{\raggedright\arraybackslash}p{4.6cm}}
  \toprule
    \textbf{Variable phase} & \textbf{Pending (dynamic)} & \textbf{Captured (lexical)} \\
  \midrule
    Meaning   & a not-yet-bound free variable & an already-obtained value, captured by a closure \\
    Mechanism & resolution of the requirement by the call context & type-level leak control \\
  \bottomrule
  \end{tabular}
\end{table}

A single family of effect systems may engage both phases at once. Thus, in the scenario
from Section~\ref{sec:es-intro}, the value for \lstinline|?db| is provided by the call
site (dynamic binding); the obtained value may be fixed by a closure as a lexical free
variable (capture), and then escape analysis must retain it within the region of the
corresponding context.

In the following sections we use this optic to take a fresh look at existing families of
effect systems; the constructive upshot --- the assembly of the static discipline of an
effect system from the mechanisms of the two phases of the cycle --- is discussed in
Section~\ref{subsec:synthesis}.

Let us reiterate: we offer a unifying point of view, not a formal classification and not a
proof of the expressive equivalence of all systems. We stress: the lifecycle describes not
which states a variable may occupy \enquote{in reality}, but which of its \emph{transitions}
a given effect-system \emph{discipline} supports and how it reflects them in the type. In
this work we focus on effects of interaction with the execution context, where the reading
through contextual free variables is most precise.


\section{Row-Based Effect Systems}
\label{sec:rows}

Most effect systems represent a set of effects by \emph{rows}~\cite{gaster1996polymorphic}.
This section considers classical row-based systems, distinguished by storing in the rows
information about \emph{all} effects of functions. Labels are the names of effects
(possibly with type arguments), and a row is an unordered set of labels; we will write it
after the result type, following a slash. Returning to the running example from
Section~\ref{sec:es-intro}: access to the database becomes the effect \lstinline|db|, and
its label in the row declares the requirement of the function body:
\begin{lstlisting}[language=colang]
    fun business(): Result / {db} { query("...") }
\end{lstlisting}
The operation \lstinline|query|, like the variable \lstinline|?db| before, will be bound
to an implementation by a handler from the dynamic call context. We stress: to enumerate
such pending variables, row-based systems need special row types on the arrows.

Higher-order functions produce at least the same effects as their arguments. Row-based
systems express this by \emph{row polymorphism}~\cite{gaster1996polymorphic}: type
variables may denote rows and be spliced into other rows. The vertical bar extends a row by
an effect variable \texttt{e}:
\begin{lstlisting}[language=colang]
    fun printMap<e, a, b>(f: (a) => b / {e},
                          xs: List<a>): List<b> / {print | e}
\end{lstlisting}
In general, each higher-order function must be parameterized by the effects of its
functional arguments, which substantially hinders a smooth migration to such an effect
system and introduces significant syntactic noise; effect inference (for example, in Koka
and Flix) mitigates this burden but does not eliminate rows from public signatures.

An example of a row-based system is Java's checked exceptions~\cite{gosling2000java}: the
language statically verifies that all checked exceptions are caught or explicitly declared
in the method signature. This feature has gained a notorious reputation among
programmers~\cite{checked-exceptions}: the exception row is an ad hoc construct that cannot
be used to parameterize a function, so it is impossible to write a higher-order method
that forwards an arbitrary number of checked exceptions.

When rows are not supported directly, they are emulated by other means. The \texttt{mtl}
library in Haskell\footnote{\url{https://hackage.haskell.org/package/mtl}} enumerates
effects by type-class constraints~\cite{jones1995functional}: being unordered and
supporting subtyping, they act as ad hoc rows. Let us note that only the recording of
requirements in the type is row-based here; binding is resolved statically, by dictionary
passing --- in essence, by implicit parameters (cf. Section~\ref{sec:capabilities}).
Haskell effect-handler libraries also use type classes for open
unions~\cite{swierstra2008data}; another approach is to emulate rows by intersection
types~\cite{xie2020row}, a practical example being the ZIO
library\footnote{\url{https://zio.dev/}} for Scala.

Classical row-based systems are the most straightforward approach, so new languages often
take it as a basis: the effect systems of Koka~\cite{leijen2014koka} and
Links~\cite{hillerstrom2016liberating} are built around rows; in Flix the inference of
polymorphic effects relies on Boolean unification~\cite{madsen2020polymorphic}, and the
Unison language realizes the row discipline in the form of
\emph{abilities}\footnote{The term \enquote{abilities} in Unison and Frank denotes the
row-based recording of the effect context and does not coincide with
\enquote{capabilities} --- the witness values of
Section~\ref{sec:capabilities}.}~\cite{unison-abilities}. However, such systems track
\emph{too much} information, leading to excessive verbosity.

From the standpoint of the proposed view, a row-based system makes the \emph{pending}
phase explicit. The effect row on an arrow is the type-level trace of the dynamic variables
of the function body that are not yet bound: each label \texttt{db}, \texttt{print}, or
\texttt{exn} denotes an operation for which a handler must be found in the execution
context. In this sense the row plays the role of a list of free dependencies, although the
labels themselves are not term variables but the names of the corresponding requirements
to the context. A handler, having bound an operation to an implementation, \emph{subtracts}
its label from the computation's row --- this is how effect safety is ensured
(Section~\ref{sec:es-intro}).

No less important is what this discipline does \emph{not} allow: binding a dynamic variable
with a value and then capturing it --- such a capture is inexpressible in the type
discipline itself. The binding of an operation to a handler is resolved anew at each
invocation, so the type of a deferred computation continues to enumerate the variables as
pending, into whatever context it may fall. Two consequences follow. First, row-based
systems do not need escape analysis: bound values are stored nowhere, and there is nothing
to leak. Second, this is precisely the root of verbosity: a dynamic variable cannot
\enquote{hide} inside a value, so it has to be repeated in the type of every intermediate
function --- hence the pervasive row polymorphism and the unavailability of contextual
polymorphism.

\enquote{Dynamism} here (see Section~\ref{sec:idea}) is about the spirit of the discipline
(binding is resolved by the calling context), whereas the checking of requirements itself
remains \emph{static}; it is precisely this that distinguishes row-based systems from
untyped dynamic binding.

The convergence of rows with the dynamic discipline also has known limits. First, there
exist row-based systems with \emph{lexically} bound handlers: the association of an
operation with a handler is fixed at the point of definition, rather than by lookup in the
dynamic call context~\cite{biernacki2019binders}; handler tunneling likewise combines the
row-based recording with a lexical discipline, solving the problem of accidental
handling~\cite{zhang2016accepting, zhang2020handling}. Second, the technique of evidence
passing compiles Koka's row-based effects into an explicit passing of handler witness
values~\cite{xie2021generalized} --- that is, pending requirements are systematically
transformed into fulfilled ones, obtained as parameters. Both observations do not refute
the thesis but refine it. The first shows that the row-based recording of a requirement is
separable from the discipline of binding: the row fixes only the pending state, while it
may be resolved even lexically. The second --- that the phase itself is separable: the
translation preserves the variable itself, changing only its representation --- from a
pending label to an obtained witness. Hence the columns of Table~\ref{tbl:free-vars} are
two representations of one and the same contextual variable, and a family chooses which of
them serves as primary at the type level. The difference is in the representation, not in
what is tracked; but, as Section~\ref{sec:comparison} will show, this choice has a real
price: escape analysis, the kind of polymorphism, verbosity.


\section{Capability-Based Effect Systems}
\label{sec:capabilities}

Earlier we saw effects as calls to \enquote{special} functions and enumerated operation
labels in row-based types. Now let us look at effect tracking differently: let effects be
performed only through \enquote{special} objects --- \emph{capabilities}, witness values
that the context supports the corresponding effect; then the system can track these objects
instead of the operations themselves.

Let us return to the running example. Explicit passing of the connection as an argument
(\lstinline|business2|) is statically safe but noisy; the dynamic variable \lstinline|?db|
is convenient but unsafe. To combine the advantages of both, we approximate a dynamically
bound variable by an \emph{implicit parameter} of the function~\cite{lewis2000implicit}: we
declare a group of such parameters by the keyword \lstinline|context|; a
\lstinline|context let| declaration at the call site makes the value available to the
compiler for automatic substitution.
\begin{lstlisting}[language=colang]
    context(db: Db)
    fun business4() { db.query("...") }

    fun caller() {
        context let db = Db.open(...)
        business4()
    }
\end{lstlisting}
Implicit parameters type the contextual free variables of a function: to call such a
function, the context must provide concrete values --- those very capabilities. The list of
a computation's implicit parameters is precisely the list of its pending variables, or, in
technical terms, of its contextual requirements: the pending phase here is also explicit,
but is expressed by the very presence of implicit parameters on the computation, and no
special row types are needed to record it. Exceptions and other algebraic operations are
tracked analogously~\cite{odersky2021safer}: the construct \lstinline|throw SomeException()|
can be understood as a call to a function that requires an implicit parameter of type
\lstinline|CanThrow<SomeException>|, while the constructs that handle exceptions provide
suitable witnesses.

The main new circumstance, however, is that capabilities can be captured by first-class
functions. Thus arises a lightweight alternative to classical effect polymorphism --- the
\emph{contextual polymorphism} promised in
Section~\ref{sec:es-intro}~\cite{brachthauser2022effects}: the \texttt{map} function can be
given its usual type without any explicit effect polymorphism:
\begin{lstlisting}[language=colang]
    fun map<a, b>(f: (a) => b, xs: List<a>): List<b>
\end{lstlisting}

Capability-based systems can be emulated even without built-in support, in a certain
programming style: functions do not use global definitions for effects but instead accept
capability objects and call their methods. The idea is realizable in Scala with the help of
implicit parameters~\cite{odersky2004scala}; the language is evolving toward built-in
exception checking with the \enquote{magic} capability
\lstinline[language=scala]|CanThrow|~\cite{odersky2021safer}.

Unfortunately, ensuring the safety of capability-based systems requires additional effort:
a capability is easily captured by a closure and leaks out of its context. For example, if
a \lstinline|CanThrow| witness leaves the bounds of a \lstinline|try-catch| block, the
guarantee that all exceptions are handled --- and with it effect safety
(Section~\ref{sec:es-intro}) --- will be lost:
\begin{lstlisting}[language=colang]
    fun unsafe() {
        var f
        try { // capability h: CanThrow<SomeException> is available
            let g = () => throw SomeException() // captures h
            f = g // leak of a computation that captured the capability
        } catch (e: SomeException) { ... }
        f() // throws SomeException outside the handler
    }
\end{lstlisting}
To prevent a leak, escape analysis at the type level is needed: its results are reflected
directly in the signatures --- which arguments are used only in place, and which may be
embedded into the result. The idea of making the lifetime of a value an object of typing
goes back to the region calculus of Tofte and Talpin~\cite{tofte1997region}. Let us add an
observation: a file descriptor, being in essence a capability to work with a file, becomes
invalid over time. The closeness of the notions of resource and capability is apparent:
both have a limited lifetime.

In the proposed optic, capability-based systems make \emph{both} phases of the cycle
explicit. The pending phase, as already noted, is expressed by the list of a computation's
implicit parameters --- it plays a role close to an effect row, but without special row
types. What is new, however, is the explicit \emph{capture} phase --- exactly the one
unavailable to row-based systems: the value of a dynamic variable here can be captured by a
closure and carried inside the computation. Dynamic binding and capture often turn out to
be a single step: a capability is \emph{obtained} from the context by an implicit parameter
and becomes a lexically visible free variable of a closure
(Table~\ref{tbl:free-vars}).

Two consequences follow that explain the makeup of the family. First, contextual
polymorphism is a direct consequence of capturing lexically visible capabilities: a
higher-order function need not be polymorphic in effects, because its argument already
carries its own capabilities. Second, the main difficulty becomes not the accounting of
pending variables but \emph{escape analysis}: one must guarantee that a value that has
captured a capability does not outlive the region of the handler or other context that
provided that capability.

The prohibition of leakage is realized in various ways. Second-class values forbid
capabilities from entering closures, results, and mutable
cells~\cite{osvald2016gentrification}. The Effekt language relaxes the prohibition by an
explicit \emph{boxing}: \texttt{box} lifts a term's captured capabilities into its type,
and \texttt{unbox} checks their availability at the point of
use~\cite{hannan1998type, brachthauser2022effects}. Capture checking in Scala assigns to
each value a \emph{capture set} and tracks it under
subtyping~\cite{boruch2023capturing}. All three mechanisms are different answers to one and
the same question of leak control, differing in price: the rigidity of the split into
classes, boxing annotations, or capture sets in the types. A fuller discussion can be found
in our earlier paper~\cite{stoyan2024modern}.


\section{Modal Effect Systems}
\label{sec:modal}

Capability-based systems made it possible to describe only part of the surrounding context
--- only those capabilities directly required by the computation. However, the price was
retention: each of the leak-prohibition mechanisms has its own cost
(Section~\ref{sec:capabilities}), growing in the limit to a dependence of types on
terms~\cite{boruch2023capturing}. This section considers a system that also strives to
specify explicitly as few parts of the context as possible, but answers leakage
differently, remaining entirely at the type level.

Originally, the Frank language proposed \emph{abilities} and \emph{adjustments} ---
second-class type-level constructs (not to be confused with the second-class values of
Section~\ref{sec:capabilities}) as syntactic sugar for row
polymorphism~\cite{lindley2017do, convent2020doo}. Later, a self-standing modal system was
proposed that makes these constructs first-class~\cite{tang2025modal}, following the idea
of using contextual modal type theory~\cite{nanevski2008contextual} to track algebraic
effects~\cite{zyuzin2021contextual}. The key idea: modality constructors should mark the
\emph{transitions} between effect contexts, rather than repeat the entire context.

The system uses two modality constructors, parameterized by effect labels; their values are
first-class (they may be stored and passed freely). The general idea is to box computations
and ascribe to them a modality describing the required context; in a suitable context the
computations are automatically unboxed. The \emph{absolute} modality (square brackets)
describes the entire required context. For example, \texttt{map} imposes no constraints on
the context, while the familiar \texttt{business} function requires the effect \texttt{db}
from Section~\ref{sec:rows}:
\begin{lstlisting}[language=modal]
    map      : []((a -> b) -> List a -> List b)
    business : [db](1 -> Result)
\end{lstlisting}
The empty brackets \texttt{[]} in the type of \texttt{map} say that the body is closed and
does not depend on the effect context, so it can be moved anywhere; the \texttt{[db]} on
\texttt{business} --- that running it requires a context handling \texttt{db}; the type
\texttt{1} here denotes the unit type.

The \emph{relative} modality (angle brackets) describes relative requirements: a boxed
computation may use both the effects of the surrounding context and those listed in the
type. Its natural habitat is the types of \emph{effect handlers}, which we have not written
in code until now. Let us introduce them into our example language: the effect
\lstinline|yield| declares a single operation, and the handler in the function
\lstinline|asList| delimits the region where this operation is available and interprets it,
accumulating elements into a list:
\begin{lstlisting}[language=colang]
    effect yield<a> { op yield(x: a): Unit }

    fun asList(f: () => Unit / {yield<Int>}): List<Int> {
        var acc = []
        handle (f()) { op yield(x) { acc = acc ++ [x]; resume(()) } }
        return acc
    }
\end{lstlisting}
The computation \lstinline|f| acts as a client: an invocation of \lstinline|yield| transfers
control to the handler, which appends an element and resumes the client by a call to
\lstinline|resume|.

In terms of the lifecycle of a contextual free variable, an invocation of \lstinline|yield|
(an effectful operation) is an \emph{occurrence} of such a variable, free in the lexical
sense, while the handler \emph{binds it dynamically}: it resolves this occurrence within its
region, just as the calling context resolves \lstinline|?db|. What in the capability family
looks like \enquote{the handler provides a capability, and the operation awaits it} is a
reification of this binding as a witness value; the row-based and modal families resolve the
same occurrence without reifying it as a passed value.

In the modal system, the type of the argument of the handler \lstinline|asList| is recorded
with a relative modality:
\begin{lstlisting}[language=modal]
    asList : <yield>(1 -> 1) -> List Int
\end{lstlisting}
The angle brackets \texttt{<yield>} mean that the argument may perform \texttt{yield}
\emph{in addition} to the effects of the surrounding context, while \texttt{asList} itself
handles this \texttt{yield}.

There is a rule: a variable should be used in a context compatible with the one in which it
was bound; it ensures both safety and encapsulation of
effects~\cite{lindley2018encapsulating}. The resulting system is expressive enough to encode
the programs of row-based systems~\cite{tang2025modal}, provided that the effect variables
on arrows refer to the lexically nearest enclosing effect context. However, it still
requires many explicit annotations --- boxing, masking~\cite{stoyan2024modern}, and a number
of technical marks~\cite{emrich2020freezeml} --- and the kind system further complicates the
picture.

From the standpoint of our reading, the modal system works with the same capture phase as
capability-based systems and in essence differs only in its answer to leakage: instead of
simply forbidding a captured contextual variable to leave the context, it re-exposes it as a
requirement to the future context --- traversing the backward edge of
Fig.~\ref{fig:lifecycle}. This works at the junction of two notions. The modality ascribed to
a boxed computation expresses its \emph{residual pending} requirement to the context: in a
suitable context the computation is automatically unboxed and becomes ready to run. The kind
system, in turn, tracks \emph{captured} (lexical) dependencies: it distinguishes values known
to be independent of the surrounding effect context (kind $Abs$ --- such are the values of
absolute modalities, freely transferable between contexts) from values that may carry a
captured dependency (kind $Any$)~\cite{tang2025modal}. When a value of kind $Any$ --- for
example, a function defined within the region of the \texttt{yield} handler and using this
effect as a lexically visible free variable --- is transferred to another context (say,
returned from the handler), the system does not necessarily reject the program: the
dependency may be expressed in the type by a relative modality that makes it explicit again.
We call this transition the \emph{uncapturing} of a variable: a bound contextual variable,
captured by a closure, is moved back into the pending phase and thereafter tracked in the
type as a contextual requirement. The term fixes the abstraction proposed here over the
mechanism of modal types, not a new rule on top of the system~\cite{tang2025modal}.

Let us demonstrate uncapturing on a familiar effect. Suppose that within the region of the
\texttt{yield} handler a function \texttt{gen'} is defined that performs this effect. As long
as it is used there, its type is the ordinary \texttt{1 -> 1}, and the dependency on
\texttt{yield} remains a captured lexical variable. But if \texttt{gen'} leaves the handler's
region (for example, is returned as a result), the type is augmented with a relative
modality:
\begin{lstlisting}[language=modal]
    gen' : 1 -> 1           // inside the yield handler's region
    gen' : <yield>(1 -> 1)  // after transfer beyond it
\end{lstlisting}
The captured variable has become pending again and visible in the type: \texttt{gen'} can now
be applied only in a context that supports \texttt{yield}.

Thus, here leak control does not reduce to a \emph{prohibition}, as in
Section~\ref{sec:capabilities}: part of the leaks turns into the re-exposure of a contextual
variable in the type. We stress that the prevention of unsafe leaks does not disappear: it is
ensured by the rule of compatibility of the contexts of use and binding of a variable together
with the kind system; the difference from capability-based systems is that this mechanism does
not form a separate subsystem with its own annotations but is built into the discipline of
modalities.


\section{Comparative Analysis}
\label{sec:comparison}

Let us bring the three families into a single picture. The proposed point of view reduces the
comparison to two questions about the lifecycle of a contextual free variable: does the system
allow it to be bound with a value from the context and that value to be captured, and, if it
does, what happens upon its leakage from the context. Row-based systems answer \enquote{no} to
the first question: variables always remain in the pending phase. Capability-based and modal
systems answer \enquote{yes} and diverge on the second: the former forbid leakage, the latter
return the leaked variable to the pending phase. The recording of the pending phase itself also
differs: row-based systems need special row types to enumerate the variables, whereas in
capability-based systems it is expressed by the very list of a computation's implicit
parameters. Everything else --- what exactly is tracked, what polymorphism is required, and, in
particular, where the annotations concentrate --- turns out to be a \emph{consequence} of these
answers, not an independent dimension. The comparison is given in Table~\ref{tbl:compare}.

\begin{table}[t]
  \caption{Three families of effect systems along a single axis: the lifecycle phases and
  transitions made explicit}
  \label{tbl:compare}
  \centering
  \footnotesize
  \begin{tabular}{>{\raggedright\arraybackslash}p{2.6cm}>{\raggedright\arraybackslash}p{3.2cm}>{\raggedright\arraybackslash}p{3.2cm}>{\raggedright\arraybackslash}p{3.6cm}}
  \toprule
  & \textbf{Row-based} & \textbf{Capability-based} & \textbf{Modal} \\
  \midrule
  Explicit phases and transitions & pending (row of labels) & pending (implicit parameters) and capture & pending (modalities), capture (kinds), and the backward transition to pending \\
  What is tracked & operation labels in the row & capability objects & transitions between contexts \\
  Polymorphism & row & contextual & implicit, by the enclosing context (no effect variables) \\
  Annotations (a consequence of the choice) & in the public signatures of higher-order functions & at the boundaries of capability leakage & at the points of transfer between contexts \\
  Ensuring safety & capture inexpressible; handling subtracts the label & leak prohibition by escape analysis & context-compatibility rule; a leaked variable is returned to pending \\
  Price & effect polymorphism & escape-analysis complexity & annotations, the kind system \\
  Example systems & Koka, Links, Java's checked exceptions, the mtl idiom (row-based recording) & Effekt, capture checking in Scala & modal effect types~\cite{tang2025modal} (a development of Frank's ideas) \\
  \bottomrule
  \end{tabular}
\end{table}

The pattern is visible in the consequence rows of Table~\ref{tbl:compare}: the choice of phases
and transitions predetermines both the kind of polymorphism and the places where annotations
concentrate. The pending phase spares one escape analysis but requires pervasive row
polymorphism; capture gives contextual polymorphism and dispenses with explicit effect
variables but shifts the entire burden onto retention --- by prohibiting leakage (second-class
values, boxing, capture sets) or by returning the leaked variable to the pending phase through
uncapturing.

The concreteness of the difference is convenient to see on the same function \texttt{printMap}
from Section~\ref{sec:rows}, which prints and applies a functional argument
(Table~\ref{tbl:map-three}; the first two signatures are given in our example language, the
third --- in the syntax of the modal system~\cite{tang2025modal}). In the row-based system, the
type carries both its own label \texttt{print} and the effect variable \texttt{e} of the
argument. In the capability-based system, the own effect has become an implicit parameter,
while about the effects of \lstinline|f| the type is silent thanks to contextual polymorphism;
the silence is paid for by the annotation \lstinline|local| --- the verdict of escape analysis
directly in the signature: the argument is used only in place. In the modal system, the
requirement \texttt{print} is expressed by an absolute modality: as with \texttt{business}, it
is pending \enquote{from birth}. About the effects of \lstinline|f| the type is silent here too:
the value is unboxed in any enclosing context containing \texttt{print}, and the inner arrows
are read relative to the context of use --- the argument may perform any of its
effects~\cite{tang2025modal} (the modality spans the whole signature; boxing the type of the
argument itself would, on the contrary, fix its effects to exactly \texttt{print}). No effect
variables are needed, and the place of the leak annotation is taken by uncapturing: a dependency
carried out of its context is re-exposed in the type by a relative modality --- as with
\texttt{gen'} (Section~\ref{sec:modal}).

\begin{table}[t]
\caption{The type of \texttt{printMap} across the three families}
\label{tbl:map-three}
\centering
\footnotesize
\begin{tabular}{@{}l@{\hspace{6pt}}l@{}}
\toprule
\textbf{Family} & \textbf{Type of \texttt{printMap}} \\
\midrule
Row-based & \lstinline[language=colang]|fun printMap<e, a, b>(f: (a) => b / {e},| \\
          & \lstinline[language=colang]!    xs: List<a>): List<b> / {print | e}! \\
Capability-based & \lstinline[language=colang]|context(print: Print)| \\
                 & \lstinline[language=colang]|fun printMap<a, b>(local f: (a) => b, xs: List<a>): List<b>| \\
Modal            & \lstinline[language=modal]|printMap : [print]((a -> b) -> List a -> List b)| \\
\bottomrule
\end{tabular}
\end{table}


\section{Synthesis: An Effect System from Two Familiar Parts}
\label{subsec:synthesis}

A single axis suggests a constructive programme: the static discipline of an effect system,
understood as the management of contextual free variables, can be designed as a combination of
two independently useful and already practitioner-familiar mechanisms --- \emph{implicit
parameters} (for the pending phase) and \emph{type-level escape analysis} (for the capture
phase). The recipe mirrors the lifecycle (Fig.~\ref{fig:lifecycle}):
\begin{enumerate}
    \item pending requirements are declared as implicit parameters: the function
    \lstinline|business4| from Section~\ref{sec:capabilities} does not open the connection
    itself and does not drag it along as an extra argument, but declares
    \lstinline|context(db: Db)|;
    \item binding is performed by the call site, providing the value through
    \lstinline|context let|;
    \item a closure may capture the obtained value --- this gives contextual polymorphism ---
    while escape analysis retains the captured one within the region of the context, and its
    verdict is visible directly in the signature (the annotation \lstinline|local| in
    Table~\ref{tbl:map-three});
    \item an attempt to carry the value across the context boundary is met either by a
    prohibition (Section~\ref{sec:capabilities}) or by uncapturing (Section~\ref{sec:modal});
    note that in the second case, too, determining the place in the program where the release
    must be performed requires some form of escape analysis.
\end{enumerate}

Implicit parameters are easier to adopt in mainstream languages than effect rows: no special
row types are needed, and their implementations already
exist~\cite{KEEP-context-parameters, oliveira2010type, lewis2000implicit}. The usefulness of
type-level escape analysis, in turn, extends beyond effect systems: it makes it possible to
reduce interprocedurally the number of heap allocations~\cite{lorenzen2024oxidizing}, to
support such language mechanisms as ownership-based borrowing in resource
management~\cite{matsakis2014rust, lorenzen2024oxidizing} and non-local
returns~\cite{akhin2021kotlin}.


\section{Conclusion}
\label{sec:conclusion}

We have shown that static control over effects is important for the reliability of programs and
for building clear abstraction boundaries. The main thesis of the work is that three seemingly
unrelated families of effect systems --- row-based, capability-based, and modal --- are variants
of a single mechanism: the management of the contextual free variables of a computation at the
type level. The families differ in which transitions of such a variable's lifecycle --- from
pending to captured and back --- they make expressible in types, and in what happens upon the
leakage of a captured variable (Table~\ref{tbl:compare}).

This unified view not only orders existing constructions but also suggests the structure of the
design space. We hope that this will help researchers and programming-language developers choose
a solution matching their tasks, and will also accelerate the adoption of effect systems into
mainstream programming languages.

A natural continuation of the work is a formalization of the proposed view: a precise
formulation of the translations between phases (in the spirit of evidence
passing~\cite{xie2021generalized}) and a metatheory of a minimal language assembled from
implicit parameters and type-level escape analysis.


\printbibliography


\section*{About the author}
\noindent Andrey Sergeevich \textsc{Stoyan} --- PhD student at the Department of Informatics,
National Research University Higher School of Economics (Saint Petersburg). Research interests:
programming language design, type systems, effect systems.

\end{document}
