% AUTO-GENERATED from bib.bib by gen_refs.py (citation order); edit the .bib, not this file.
\begin{thebibliographyEng}{99}

\bibitem{park1992escape}
Park Y.G., Goldberg B. Escape analysis on lists. Proceedings of the ACM SIGPLAN 1992 conference on Programming language design and implementation. 1992. pp. 116--127.

\bibitem{blanchet1999escape}
Blanchet B. Escape analysis for object-oriented languages: application to Java. \emph{Acm Sigplan Notices.} 1999. vol. 34. no. 10. pp. 20--34.

\bibitem{matsakis2014rust}
Matsakis N.D., Klock F.S. The rust language. Proceedings of the 2014 ACM SIGAda annual conference on High integrity language technology. 2014. pp. 103--104.

\bibitem{lorenzen2024oxidizing}
Lorenzen A., White L., Dolan S., Eisenberg R.A., Lindley S. Oxidizing OCaml with modal memory management. \emph{Proceedings of the ACM on Programming Languages.} 2024. vol. 8. no. ICFP. pp. 485--514.

\bibitem{georges2025data}
Georges A.L., Peters B., Elbeheiry L., White L., Dolan S., Eisenberg R.A., Casinghino C., Pottier F., Dreyer D. Data Race Freedom {\`a} la Mode. \emph{Proceedings of the ACM on Programming Languages.} 2025. vol. 9. no. POPL.

\bibitem{akhin2021kotlin}
Akhin M., Belyaev M. Kotlin language specification. \emph{Kotlin Language Specification.} 2021.

\bibitem{jung2018rustbelt}
Jung R., Jourdan J., Krebbers R., Dreyer D. RustBelt: Securing the Foundations of the Rust Programming Language. \emph{Proceedings of the ACM on Programming Languages.} 2018. vol. 2. no. POPL. pp. 1--34.

\bibitem{jung2020stacked}
Jung R., Dang H., Kang J., Dreyer D. Stacked Borrows: An Aliasing Model for Rust. \emph{Proceedings of the ACM on Programming Languages.} 2020. vol. 4. no. POPL. pp. 1--32.

\bibitem{astrauskas2022prusti}
Astrauskas V., B{\'{i}}l{\'{y}} A., Fiala J., Grannan Z., Matheja C., M{\"{u}}ller P., Poli F., Summers A.J. The Prusti Project: Formal Verification for Rust. NASA Formal Methods - 14th International Symposium, NFM 2022. 2022. pp. 88--108.

\bibitem{denis2022creusot}
Denis X., Jourdan J., March{\'{e}} C. Creusot: A Foundry for the Deductive Verification of Rust Programs. Formal Methods and Software Engineering - 23rd International Conference on Formal Engineering Methods, ICFEM 2022. 2022. pp. 90--105.

\bibitem{matsushita2022rusthornbelt}
Matsushita Y., Denis X., Jourdan J., Dreyer D. RustHornBelt: a semantic foundation for functional verification of Rust programs with unsafe code. PLDI '22: 43rd ACM SIGPLAN International Conference on Programming Language Design and Implementation. 2022. pp. 841--856.

\bibitem{gaher2024refinedrust}
G{\"{a}}her L., Sammler M., Jung R., Krebbers R., Dreyer D. RefinedRust: A Type System for High-Assurance Verification of Rust Programs. \emph{Proceedings of the ACM on Programming Languages.} 2024. vol. 8. no. PLDI. pp. 1115--1139.

\bibitem{plotkin2013handling}
Plotkin G.D., Pretnar M. Handling algebraic effects. \emph{Logical methods in computer science.} 2013. vol. 9.

\bibitem{brachthauser2020effects}
Brachth{\"a}user J.I., Schuster P., Ostermann K. Effects as capabilities: effect handlers and lightweight effect polymorphism. \emph{Proceedings of the ACM on Programming Languages.} 2020. vol. 4. no. OOPSLA. pp. 1--30.

\bibitem{brachthauser2022effects}
Brachth{\"a}user J.I., Schuster P., Lee E., Boruch-Gruszecki A. Effects, capabilities, and boxes: from scope-based reasoning to type-based reasoning and back. \emph{Proceedings of the ACM on Programming Languages.} 2022. vol. 6. no. OOPSLA1. pp. 1--30.

\bibitem{boruch2023capturing}
Boruch-Gruszecki A., Odersky M., Lee E., Lhot{\'a}k O., Brachth{\"a}user J. Capturing types. \emph{ACM Transactions on Programming Languages and Systems.} 2023. vol. 45. no. 4. pp. 1--52.

\bibitem{xie2021generalized}
Xie N., Leijen D. Generalized evidence passing for effect handlers: efficient compilation of effect handlers to C. \emph{Proceedings of the ACM on Programming Languages.} 2021. vol. 5. no. ICFP. pp. 1--30.

\bibitem{xie2022first}
Xie N., Cong Y., Ikemori K., Leijen D. First-class names for effect handlers. \emph{Proceedings of the ACM on Programming Languages.} 2022. vol. 6. no. OOPSLA2. pp. 30--59.

\bibitem{stoyan2025existential}
Stoyan A.S. Type-based Escape Analysis with Existential Lifetimes. \emph{Trudy ISP RAN / Proc. ISP RAS.} 2026. vol. 38. no. 1. pp. 71--78.

\bibitem{tofte1997region}
Tofte M., Talpin J. Region-Based Memory Management. \emph{Information and Computation.} 1997. vol. 132. no. 2. pp. 109--176.

\bibitem{fluet2006monadic}
Fluet M., Morrisett G. Monadic Regions. \emph{Journal of Functional Programming.} 2006. vol. 16. no. 4--5. pp. 485--545.

\bibitem{zyuzin2021contextual}
Zyuzin N., Nanevski A. Contextual modal types for algebraic effects and handlers. \emph{Proceedings of the ACM on Programming Languages.} 2021. vol. 5. no. ICFP. pp. 1--29.

\bibitem{van2022handling}
van der Rest C., Reinders J., Poulsen C.B. Handling Higher-Order Effects. \emph{arXiv preprint arXiv:2203.03288.} 2022.

\bibitem{van2024framework}
van den Berg B., Schrijvers T. A framework for higher-order effects \& handlers. \emph{Science of Computer Programming.} 2024. vol. 234. pp. 103086.

\bibitem{tang2025modal}
Tang W., White L., Dolan S., Hillerstr{\"o}m D., Lindley S., Lorenzen A. Modal effect types. \emph{Proceedings of the ACM on Programming Languages.} 2025. vol. 9. no. OOPSLA1. pp. 1130--1157.

\bibitem{sivaramakrishnan2021retrofitting}
Sivaramakrishnan K.C., Dolan S., White L., Kelly T., Jaffer S., Madhavapeddy A. Retrofitting effect handlers onto OCaml. PLDI '21: 42nd ACM SIGPLAN International Conference on Programming Language Design and Implementation. 2021. pp. 206--221.

\bibitem{tang2024soundly}
Tang W., Hillerstr{\"o}m D., Lindley S., Morris J.G. Soundly Handling Linearity. \emph{Proceedings of the ACM on Programming Languages.} 2024. vol. 8. no. POPL.

\bibitem{vanrooij2025affect}
van Rooij O., Krebbers R. Affect: An Affine Type and Effect System. \emph{Proceedings of the ACM on Programming Languages.} 2025. vol. 9. no. POPL.

\bibitem{wu2014effect}
Wu N., Schrijvers T., Hinze R. Effect handlers in scope. Proceedings of the 2014 ACM SIGPLAN Symposium on Haskell. 2014. pp. 1--12.

\bibitem{biernacki2019binders}
Biernacki D., Pir{\'o}g M., Polesiuk P., Sieczkowski F. Binders by day, labels by night: effect instances via lexically scoped handlers. \emph{Proceedings of the ACM on Programming Languages.} 2019. vol. 4. no. POPL. pp. 1--29.

\bibitem{odersky2021safer}
Odersky M., Boruch-Gruszecki A., Brachth{\"a}user J.I., Lee E., Lhot{\'a}k O. Safer exceptions for Scala. Proceedings of the 12th ACM SIGPLAN International Symposium on Scala. 2021. pp. 1--11.

\bibitem{odersky2022scoped}
Odersky M., Boruch-Gruszecki A., Lee E., Brachth{\"a}user J., Lhot{\'a}k O. Scoped capabilities for polymorphic effects. \emph{arXiv preprint arXiv:2207.03402.} 2022.

\bibitem{KEEP-context-parameters}
Serrano A., Akhin M., Bobko N., Gorbunov I., Zarechenskii M., Zharkov D. Context parameters (KEEP-367). 2024. Available at: \url{}.

\bibitem{osvald2016gentrification}
Osvald L., Essertel G., Wu X., Alay{\'o}n L.I.G., Rompf T. Gentrification gone too far? affordable 2nd-class values for fun and (co-) effect. \emph{ACM SIGPLAN Notices.} 2016. vol. 51. no. 10. pp. 234--251.

\bibitem{leijen2017type}
Leijen D. Type directed compilation of row-typed algebraic effects. Proceedings of the 44th ACM SIGPLAN Symposium on Principles of Programming Languages. 2017. pp. 486--499.

\bibitem{zhang2016accepting}
Zhang Y., Salvaneschi G., Beightol Q., Liskov B., Myers A.C. Accepting blame for safe tunneled exceptions. \emph{ACM SIGPLAN Notices.} 2016. vol. 51. no. 6. pp. 281--295.

\bibitem{zhang2020handling}
Zhang Y., Salvaneschi G., Myers A.C. Handling bidirectional control flow. \emph{Proceedings of the ACM on Programming Languages.} 2020. vol. 4. no. OOPSLA. pp. 1--30.

\bibitem{xu2023formalizing}
Xu Y., Odersky M. Formalizing Box Inference for Capture Calculus. \emph{arXiv preprint arXiv:2306.06496.} 2023.

\bibitem{xu2025s}
Xu Y., Bra{\v{c}}evac O., Pham C.N., Odersky M. What’s in the Box?. \emph{Ergonomic and Expressive Capture Tracking over Generic Data Structures (Extended Version),(Aug. 2025). doi.} 2025. vol. 10. pp. 17.

\bibitem{stoyan2024modern}
Stoyan A. Modern Effect Systems Overview. 2024 Ivannikov Ispras Open Conference (ISPRAS). 2024. pp. 1--7.

\bibitem{aydemir2005poplmark}
Aydemir B.E., Bohannon A., Fairbairn M., Foster J.N., Pierce B.C., Sewell P., Vytiniotis D., Washburn G., Weirich S., Zdancewic S. Mechanized Metatheory for the Masses: The POPLMark Challenge. Theorem Proving in Higher Order Logics. 2005. pp. 50--65.

\bibitem{aydemir2008engineering}
Aydemir B.E., Chargu{\'e}raud A., Pierce B.C., Pollack R., Weirich S. Engineering Formal Metatheory. \emph{ACM SIGPLAN Notices.} 2008. vol. 43. no. 1. pp. 3--15.

\bibitem{sewell2010ott}
Sewell P., Nardelli F.Z., Owens S., Peskine G., Ridge T., Sarkar S., Strni{\v{s}}a R. Ott: Effective Tool Support for the Working Semanticist. \emph{Journal of Functional Programming.} 2010. vol. 20. no. 1. pp. 71--122.

\bibitem{schaefer2015autosubst}
Sch{\"a}fer S., Tebbi T., Smolka G. Autosubst: Reasoning with de Bruijn Terms and Parallel Substitutions. Interactive Theorem Proving. 2015. pp. 359--374.

\bibitem{stark2019autosubst2}
Stark K., Sch{\"a}fer S., Kaiser J. Autosubst 2: Reasoning with Multi-sorted de Bruijn Terms and Vector Substitutions. Proceedings of the 8th ACM SIGPLAN International Conference on Certified Programs and Proofs. 2019. pp. 166--180.

\end{thebibliographyEng}
