%! suppress = EscapeAmpersand
%! suppress = TooLargeSection
%! suppress = MissingLabel

% Preamble — adapted to the layout of
% "Информатика и автоматизация / Informatics and Automation" (SPIIRAS, SPC RAS),
% https://ia.spcras.ru/index.php/sp/aims-and-scope
\documentclass[12pt,a4paper]{article}

% Packages
\usepackage[T2A,T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[english,russian]{babel}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{xcolor}
\usepackage{multicol}
\usepackage{proof}
\usepackage{mathtools}
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=black, citecolor=black, urlcolor=blue}

\usepackage{geometry}
% Journal page setup: A4, 2.0 cm margins, 1.5 line spacing.
\geometry{a4paper, top=20mm, bottom=20mm, left=20mm, right=20mm}

\usepackage{setspace}
\onehalfspacing

\usepackage{enumitem}
\setlist{noitemsep}

\usepackage{indentfirst}
\usepackage{booktabs}
\usepackage{array}

% Section heading style closer to the journal:
% centered, bold, sans-serif feel approximated with bfseries.
\usepackage{titlesec}
\titleformat{\section}{\centering\normalfont\large\bfseries}{\thesection.}{0.6em}{}
\titleformat{\subsection}{\normalfont\bfseries}{\thesubsection.}{0.6em}{}

\newtheorem{theorem}{Theorem}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}

% Inline helper for the UDC/DOI/affiliation lines in the header.
\newcommand{\headerline}[1]{\noindent\small #1\par}

\newcommand{\graybox}[1]{\colorbox{lightgray}{$\displaystyle #1$}}
\newcommand{\mathframebox}[1]{\framebox{$\displaystyle #1$}}
\newcommand{\vor}{~|~}
\newcommand{\ap}{~}
\newcommand{\ctx}[1]{ctx\left(#1\right)~}
\newcommand{\step}{\rightsquigarrow}
\newcommand{\local}{\top}
\newcommand{\free}{\bot}
\newcommand{\keyword}[1]{\mathbf{#1}}
\newcommand{\identation}{\text{\hspace{2em}}}

\usepackage{listings}
\lstdefinestyle{modern}{
    basicstyle=\normalfont\ttfamily\small,     % fixed-width font, smaller size
    keywordstyle=\color{blue!80!black}\bfseries,  % strong blue keywords
    stringstyle=\color{red!70!black},      % subtle red for strings
    commentstyle=\color{green!60!black}\itshape,  % italic green comments
    numberstyle=\tiny\color{gray!50},      % small, gray line numbers
    rulecolor=\color{gray!50},            % frame color
    tabsize=4,                           % tab size of 2 spaces
    showspaces=false,                   % don't show spaces as underscores
    showstringspaces=false,             % don't highlight spaces in strings
    breaklines=false,                    % automatic line breaking
    breakatwhitespace=false,             % break lines at whitespace
    captionpos=b,                       % caption at the bottom
    escapeinside={(*@}{@*)},             % escape to LaTeX inside comments
    gobble=8,      % skip 4 characters of indentation in code lines
}
\lstdefinelanguage{colang}{
    keywords=[1]{let, var, effect, perform, data, context, fun, where, if, else, throw, try, catch, local, free, op, handle, resume, class, interface, override, init, return},
    sensitive=true,
    comment=[l]{//},
    morecomment=[s]{/*}{*/},
    morestring=[b]",
    breaklines=false,
    tabsize=4,
    showstringspaces=false,
}
\lstdefinelanguage{rust}{
    keywords={
        as, break, const, continue, crate, else, enum, extern, false, fn, for, if, impl, in,
        let, loop, match, mod, move, mut, pub, ref, return, self, Self, static, struct, super,
        trait, true, type, unsafe, use, where, while, async, await, dyn
    },
    comment=[l]{//},
    morecomment=[s]{/*}{*/},
    morestring=[b]",
}
\lstset{style=modern}

% Document
\begin{document}
\selectlanguage{english}

    % ---------------------------------------------------------------------
    % Journal header block (English).
    % Layout follows the conventions of "Informatics and Automation"
    % (https://ia.spcras.ru): UDC, DOI placeholder, title, authors,
    % affiliation, abstract, keywords, acknowledgments.
    % ---------------------------------------------------------------------

    \headerline{\textbf{UDC} 004.052.42, 004.43}
    \headerline{\textbf{DOI} 10.15622/ia.XX.X.X\hfill\textit{(to be assigned by the editor)}}

    \vspace{0.5em}
    \begin{center}
        {\Large\bfseries
            A mechanized proof of type safety for an escape-analysis calculus
            with existential lifetimes and effect handlers\par}

        \vspace{0.8em}
        {\normalsize\textbf{A.\,Stoyan}\textsuperscript{1}\par}

        \vspace{0.4em}
        {\small\itshape
            \textsuperscript{1}\,National Research University Higher School of Economics\\
            (HSE University), 3A Kantemirovskaya Str., Saint Petersburg, 194100, Russian Federation\par}
    \end{center}

    \vspace{0.5em}
    \noindent\textbf{Abstract.}
    Type-based escape analysis is a static-analysis technique that lifts the question of
    whether a value outlives its lexical scope into the type system; it underpins
    stack allocation, borrowing in ownership-based languages, and safety of capability-based
    effect systems.
    A recent calculus, $Core_\Delta$, reduces the user-facing annotation burden of such systems
    through three combined design decisions: \emph{existential} lifetimes on compound data,
    an explicit \emph{minimum} operation over lifetimes, and bounded polymorphism with
    lifetime-annotated bounds.
    However, the meta-theoretic guarantees of $Core_\Delta$ have so far been justified only on paper,
    and the principal motivating application of the calculus~--- capability-based effect
    handlers~--- has been discussed only informally.
    \emph{The aim of this work} is to close both gaps with a machine-checked proof of type safety
    for a calculus that genuinely combines existential lifetimes with delimited-control effect handlers.
    \emph{Methods.}
    The $Core_\Delta$ calculus is extended with single-operation algebraic effect handlers in the style
    of capability passing.
    For the extended calculus we give a small-step operational semantics based on
    marker-indexed pure evaluation contexts that reify captured continuations as first-class
    runtime values, and we mechanize the entire system~--- syntax, semantics, subtyping, and typing~---
    in the Rocq/Coq proof assistant using pure de~Bruijn indices.
    \emph{Results.}
    We prove the classical progress and preservation theorems for the extended calculus and
    obtain type safety as a corollary.
    Going beyond type safety, we mechanize two escape-analysis guarantees that motivate the
    calculus: a \emph{capability-confinement} theorem stating that a runtime capability can never
    surface at an evaluation position outside its handler, and an \emph{operational non-escape}
    theorem stating that a value delivered at an escapable ($free$) data type can never carry a
    $local$-confined lifetime.
    The development comprises approximately $6{,}500$ lines of Rocq code organized into eight modules
    and is accompanied by more than thirty machine-checked examples, including the state, exception,
    and non-determinism effects encoded with the new handlers, all reduced inside the proof assistant.
    A comparison with related mechanized accounts of effect handlers and of lifetime/region systems
    shows that the present development is, to the best of our knowledge, the first to combine
    \emph{both} ingredients within a single mechanized soundness proof.
    \emph{Practical relevance.}
    The artifact provides a reusable, machine-checked foundation for designing type-based escape
    analyses in languages with first-class effect handlers; it is publicly available at
    \url{https://github.com/penguin-boxx/effect-proofs}.

    \vspace{0.4em}
    \noindent\textbf{Keywords:}
    type systems, escape analysis, existential lifetimes, algebraic effect handlers,
    delimited control, capability confinement, mechanized metatheory, Rocq/Coq, type safety.

    \vspace{0.4em}
    \noindent\textbf{Acknowledgements.}
    The author thanks the colleagues at HSE University for fruitful discussions on the design of
    capability-based effect systems.
    No external funding was received for this work.


    \section{Introduction}

    Escape analysis is a long-standing static-analysis technique whose purpose is to
    determine whether the lifetime of a value may exceed its defining lexical
    scope~\cite{blanchet1999escape, park1992escape}.
    Beyond classical compiler optimizations such as stack allocation and scalar
    replacement, escape analysis has in the last decade become a load-bearing
    component of several modern type systems: it justifies the safety of non-local
    returns in Kotlin~\cite{akhin2021kotlin}, underlies borrowing in
    ownership-based languages such as Rust~\cite{matsakis2014rust} and recent
    extensions of OCaml~\cite{lorenzen2024oxidizing}, and enforces
    \emph{contextual polymorphism} in capability-based effect
    systems~\cite{brachthauser2020effects, brachthauser2022effects, boruch2023capturing}.

    Integrating escape analysis directly into the type system has two well-known
    advantages over conventional data-flow approaches.
    First, escape information becomes explicit at function boundaries, so that
    inter\-procedural analysis reduces to intraprocedural rule-checking consulting
    function signatures.
    Second, no long-distance constraints are generated, which makes diagnostics
    local and predictable.
    The trade-off is that user-facing annotations rapidly accumulate: lifetime
    polymorphism, variance, and explicit \emph{outlives} constraints have to be
    declared on every data type and threaded through every signature, which
    raises the entry barrier both for compiler implementers and for end users.

    \subsection{Problem statement}

    A recent calculus, $Core_\Delta$~\cite{stoyan2025existential},
    attacks the annotation burden by changing the form of lifetime
    information available at type-level data:
    \begin{itemize}
        \item lifetimes form a two-point lattice $free \sqsubseteq local$ extended with a
            binary \emph{minimum} operation $+$, so that a function signature can
            express the data-flow join of its arguments directly;
        \item compound data types are not parameterized by lifetimes; instead the
            lifetime of a constructed value is the minimum of the lifetimes of its
            fields, and these fields' lifetimes are recovered \emph{existentially}
            on pattern matching;
        \item bounded polymorphism with lifetime-annotated bounds permits tracked
            (i.e.\ $local$) values to instantiate type parameters in a controlled way.
    \end{itemize}

    The proposal is attractive, but two questions left open by the original work
    threaten its viability for a real implementation.
    \emph{(Q1)} Is the type system actually sound?
    The interaction between subtyping over a non-trivial lifetime lattice, existential
    lifetimes introduced by data constructors, and variance-directed elimination on
    pattern matching is subtle enough that a paper-and-pencil argument is not
    sufficient evidence for adopting the calculus in a compiler.
    \emph{(Q2)} Does the system extend cleanly to its principal motivating application
    --- capability-based effect handlers?
    Handlers introduce a delimited-control operational semantics whose interaction
    with lifetime annotations, with substitution under multiple binder forms, and with
    the existential lifetimes introduced by constructors is non-obvious.

    The present paper provides a positive answer to both questions in the form of a
    fully machine-checked artifact.

    \subsection{Contributions}

    The contributions of this paper are the following.

    \begin{enumerate}
        \item We extend the $Core_\Delta$ calculus with single-operation algebraic effect
            handlers in the capability-passing style of Bra\-cht\-h\"au\-ser et
            al.~\cite{brachthauser2020effects}, encoding capabilities as ordinary data
            constructors that carry a runtime marker
            (Section~\ref{sec:background}, Section~\ref{sec:formalization}).
        \item We give a small-step operational semantics for the extended calculus
            based on evaluation contexts with marker-indexed \emph{pure} sub-contexts;
            this models the ``capture of the continuation up to the matching handler''
            of multi-prompt delimited control as a syntactic reification via a new
            $\keyword{resume}_m$ runtime form (Section~\ref{sec:formalization}).
        \item We mechanize the syntax, operational semantics, subtyping, and typing of
            the extended calculus in Rocq/Coq using pure de~Bruijn indices, and we
            prove the classical progress, preservation, and type-safety theorems
            (Section~\ref{sec:formalization}, Section~\ref{sec:mechanization}).
        \item Beyond type safety, we mechanize the two escape-analysis guarantees that
            motivate the calculus:
            a \emph{capability-confinement} theorem (a runtime capability is never even
            well-typed at an evaluation position outside its handler, hence a closed
            program never reduces to a configuration that exposes one), and an
            \emph{operational non-escape} theorem (a value delivered at an escapable
            $free$ data type provably carries no top-level $local$ lifetime).
            Both are stated as Rocq theorems rather than argued informally
            (Section~\ref{sec:formalization}).
        \item We exercise the development with more than thirty machine-checked
            examples, including reduction traces for state, exception, and
            non-determinism effects encoded with the new handlers
            (Section~\ref{sec:evaluation}).
        \item We compare the development with existing mechanizations of effect-handler
            calculi and of lifetime/region systems, showing that --- to the best of our
            knowledge --- this is the first mechanized soundness proof that genuinely
            combines an existential-lifetime escape analysis with first-class effect
            handlers (Section~\ref{sec:related-work}).
    \end{enumerate}

    The remainder of the paper is organized as follows.
    Section~\ref{sec:related-work} reviews related work on type-based escape
    analysis and on mechanizations of effect handlers.
    Section~\ref{sec:background} recalls the design of $Core_\Delta$ and motivates
    the effect-handler extension on a small example.
    Section~\ref{sec:formalization} presents the formal calculus and states the main
    meta-theorems together with their proof outlines.
    Section~\ref{sec:mechanization} describes the structure of the Rocq development,
    its representation choices, and the explicitly delimited scope of the axiomatized
    auxiliary lemmas.
    Section~\ref{sec:evaluation} evaluates the development through a series of
    worked examples.
    Section~\ref{sec:discussion} discusses limitations and threats to validity.
    Section~\ref{sec:conclusion} concludes.


    \section{State of the art} \label{sec:related-work}

    \subsection{Type-based escape analysis in production languages}

    Several modern languages embed a form of escape analysis directly into their
    type system.
    The most established case is the lifetime system of Rust~\cite{matsakis2014rust}:
    it combines lifetime polymorphism on data types, variance inference, and explicit
    outlives constraints to obtain a system that is highly expressive but, by common
    consensus, difficult both to teach and to implement.
    Notably, Rust expresses the minimum of two lifetimes only indirectly, either by
    introducing additional outlives constraints or by equating lifetimes.

    A more recent thread of work, exemplified by the OCaml \texttt{local}
    modality~\cite{lorenzen2024oxidizing}, treats locality as a modality on bindings
    rather than as a first-class component of the type.
    The resulting system is simpler than Rust's, but pays for that simplicity by
    being unable to instantiate polymorphic type parameters with types carrying
    a non-default modality.
    Kotlin's contract-based escape analysis~\cite{akhin2021kotlin} is similarly
    binding-oriented and is therefore limited to first-order escape conditions.

    \subsection{Type-based escape analysis in effect systems}

    Capability-based effect systems explicitly rely on escape analysis to enforce
    that a capability granted by \texttt{handle} does not outlive the dynamic
    extent of its handler.
    The Effekt language~\cite{brachthauser2020effects, brachthauser2022effects}
    pioneers this approach with a notion of \emph{second-class} bindings and uses
    them to guarantee the safety of contextual polymorphism.
    Xie et al.~\cite{xie2022first} adopt higher-rank polymorphism for a similar
    purpose in their first-class evidence-passing translation of effect handlers.
    Boruch-Gru\-szecki et al.~\cite{boruch2023capturing} introduce capture
    tracking in the Scala dependent type system to refine the escape information
    attached to function types.

    These works share with $Core_\Delta$ the goal of making escape information
    visible in types, but they differ on the carrier of that information:
    Effekt and capture-tracking Scala attach escape annotations to types as
    sets of captured variables, whereas $Core_\Delta$ uses a coarse-grained
    two-point lattice with an explicit minimum, which is easier to integrate with
    subtyping and considerably simpler to inhabit on the user side.

    \subsection{Mechanizations of effect-handler calculi}

    The meta-theory of algebraic effect handlers has been mechanized several
    times, most notably in Rocq/Coq and Agda variants of the
    $\lambda^{eff}$-style calculus~\cite{brachthauser2020effects, brachthauser2022effects}.
    These developments typically focus either on type safety with simple type
    systems (no escape analysis), or on the meta-theory of capture tracking on
    its own (no effect handlers).
    Our development closes this gap by handling both ingredients in a single
    mechanized soundness proof.

    \subsection{Comparison summary}

    Table~\ref{tab:comparison} summarizes how the present development compares
    with the closest pieces of related work.

    \begin{table}[h]
        \centering
        \small
        \caption{Comparison with closest related work on type-based escape analysis
            and on mechanizations of effect-handler calculi.
            ``Mech.'' = formally mechanized; ``EA'' = type-based escape analysis;
            ``Eff.'' = first-class effect handlers; ``Min'' = explicit minimum on lifetimes.}
        \label{tab:comparison}
        \begin{tabular}{@{}lcccc@{}}
            \toprule
            \textbf{System / work} & \textbf{Mech.} & \textbf{EA} & \textbf{Eff.} & \textbf{Min} \\
            \midrule
            Rust lifetimes~\cite{matsakis2014rust}             & partial & yes & no  & indirect \\
            OCaml \texttt{local}~\cite{lorenzen2024oxidizing}  & no      & yes & no  & no       \\
            Effekt~\cite{brachthauser2020effects}              & yes     & yes & yes & no       \\
            Capture tracking~\cite{boruch2023capturing}        & yes     & yes & no  & no       \\
            $Core_\Delta$ (paper)~\cite{stoyan2025existential} & no      & yes & no  & yes      \\
            \textbf{This work}                                 & \textbf{yes} & \textbf{yes} & \textbf{yes} & \textbf{yes} \\
            \bottomrule
        \end{tabular}
    \end{table}


    \section{Background and motivating example} \label{sec:background}

    \subsection{The $Core_\Delta$ fragment in brief}

    In $Core_\Delta$, every type carries a lifetime annotation drawn from the lattice
    $free \sqsubseteq local$, extended with a binary minimum $\Delta_1 + \Delta_2$ and
    with lifetime variables bound by the enclosing type scheme.
    The annotation $T\,'\!\Delta$ records the lifetime at which a value of type $T$ may
    safely be used.
    The lifetime $free$ classifies values that may escape arbitrarily; the lifetime
    $local$ classifies values that must not leave their defining scope.
    The crucial typing invariant, enforced by the rule for $\lambda$-abstraction, is that
    the result type of a function may not mention $local$.
    This prevents leakage of tracked values through return values, captured closures, or
    compound data.

    Compound data types in $Core_\Delta$ do \emph{not} carry user-visible lifetime
    parameters.
    Instead, the lifetime of a constructed value is computed as the minimum of the
    lifetimes of its arguments; when a value is pattern-matched, the lifetimes of the
    components are conservatively approximated by the lifetime of the scrutinee.
    Internally, the lifetimes that were ``forgotten'' at construction time are recovered
    as \emph{existential} lifetime variables, which the $Match$ rule re-binds with a
    fresh name bounded by the scrutinee's lifetime.
    These fresh variables must be eliminated from the type of each branch before they
    leak into the outer scope; this elimination is performed by a variance-directed
    substitution that approximates positive occurrences by the bound, negative
    occurrences by $free$, and rejects invariant occurrences.

    \subsection{Effect handlers as the motivating application}

    Capability-based effect systems use type-based escape analysis to ensure that a
    capability granted by \texttt{handle} cannot be smuggled out of the dynamic
    extent of the corresponding handler.
    A natural encoding maps a capability for effect $E$ with type arguments
    $\overline{T}$ to a data constructor $K_{cap_E}$ whose result type is
    $E\,'\!local\,\overline{T}$: the $local$ annotation expresses precisely the
    requirement that the capability must not escape.
    Operations on the effect are invoked via \texttt{perform}, which reifies the
    surrounding pure evaluation context as a resumption and transfers control to the
    operation body installed by \texttt{handle}.
    The resumption itself is then a value whose type must in turn prevent it from
    outliving the handler frame.

    Three aspects of this design make a paper-and-pencil soundness argument fragile:
    \begin{itemize}
        \item The reduction relation must locate the matching handler delimiter
            \emph{across} pure sub-contexts but must not cross other handlers.
            This standard ``operation-lookup'' rule of multi-prompt delimited
            control interacts with substitution under binders and therefore with
            lifetime annotations.
        \item The continuation captured by \texttt{perform} is reified as a value
            (the runtime form $\keyword{resume}_m\,b$), so the type system must
            assign it a function type while keeping the runtime marker $m$~---
            which has no static counterpart~--- implicit.
        \item Existential lifetimes introduced by constructors interact with the
            substitution lemma: pattern-matching a capability under a handler must
            re-bind its existential lifetimes without aliasing the marker-fresh
            names introduced by \texttt{perform}.
    \end{itemize}

    \subsection{Running example}\label{subsec:running-example}

    To make these interactions concrete, consider an exception effect $\mathit{Exc}$ with
    a single operation $\mathit{throw} : \forall \alpha.\,Unit \to \alpha$.
    The capability constructor produced by the effect declaration is
    $K_{cap_{\mathit{Exc}}} : Unit\,'\!local \to \mathit{Exc}\,'\!local$.
    A client uses the effect as follows (with $\beta$ a fresh type variable bound
    inside the operation body, instantiated at the \texttt{perform} site):
    \[
        \keyword{handle}~\mathit{Exc}\;[Int]\;\;\big(\lambda x.\,\lambda k.\,0\big)\;\;\keyword{in}\;\;
        \keyword{perform}\;c_{\mathit{Exc}}\;[Int]\;()\,+\,1
    \]
    Here the operation body $\lambda x.\,\lambda k.\,0$ discards the resumption~$k$ and
    returns the default value~$0$, a typical exception-style handler.
    Reduction proceeds as follows.
    First, $\keyword{handle}$ allocates a fresh marker $m$, installs a delimiter
    $\keyword{handler}_m$, and substitutes the runtime capability
    $\keyword{cap}_{\mathit{Exc}}^m\,[Int]\,(\lambda x.\,\lambda k.\,0)$ for the bound
    variable~$c_{\mathit{Exc}}$.
    Second, the $H\!\_Perform$ rule identifies the surrounding pure context $P_m =
    [\cdot]\,+\,1$ and rewrites the configuration to
    \[
        (\lambda x.\,\lambda k.\,0)\;()\;\;(\keyword{resume}_m\,(([\cdot]\,+\,1)\;\$0)),
    \]
    discharging the resumption.
    Third, two $\beta$-steps eliminate the operation body and yield the value~$0$,
    then $H\!\_Return$ removes the handler delimiter to produce $0$ as the final
    answer.

    Note how the $local$ annotation on the capability type ensures that even though the
    body $\lambda x.\,\lambda k.\,0$ has the resumption $k$ in scope, it cannot smuggle
    $k$ out via its closure environment, because doing so would require typing a closure
    with $local$ in a covariant position of its result type --- which the $T\!\_Lam$
    rule forbids.
    This very interaction is what the mechanization makes precise.


    \section{Formalization} \label{sec:formalization}

    \subsection{Syntax}

    Lifetimes, types, and terms follow the grammar of $Core_\Delta$, extended with three
    runtime-only term forms for the operational semantics of handlers
    (Figure~\ref{fig:syntax}).
    \begin{figure}[h]
        \centering
        \[
            \begin{array}{lrl}
                \text{Lifetimes}        & \Delta & \Coloneqq l \vor free \vor local \vor \Delta_1 + \Delta_2 \\
                \text{Types}            & \tau   & \Coloneqq \alpha \vor \overline{\tau}\,\Delta \to \sigma
                                                    \vor K\,\Delta\,\overline{\tau}
                                                    \vor \forall l.\,\tau \vor \forall (\alpha <: \tau').\,\tau \\
                \text{Terms}            & t      & \Coloneqq x \vor t\,t \vor \lambda(x : \tau).\,t
                                                    \vor t\,[\tau] \vor \Lambda(\alpha <: \tau).\,t
                                                    \vor t\,[\Delta] \vor \Lambda l.\,t \\
                                        &        & \vor K\,\Delta\,\overline{\Delta}\,\overline{\tau}\,\overline{t}
                                                    \vor \keyword{match}\,t\,\keyword{with}\,K\,\overline{x} \Rightarrow t \mid \_ \Rightarrow t \\
                                        &        & \vor \keyword{handle}\,E\,\overline{\tau}\,t_{op}\,\keyword{in}\,t
                                                    \vor \keyword{perform}\,t\,\overline{\tau}\,t \\
                                        &        & \vor \keyword{cap}_E^m\,\overline{\tau}\,t_{op}
                                                    \vor \keyword{handler}_m\,t
                                                    \vor \keyword{resume}_m\,b
            \end{array}
        \]
        \caption{Syntax of the extended calculus.}
        \label{fig:syntax}
    \end{figure}
    Markers $m \in \mathbb{N}$ are fresh natural numbers generated by \texttt{handle}.
    The forms $\keyword{cap}_E^m$, $\keyword{handler}_m$, and $\keyword{resume}_m$ never
    appear in source programs; they arise only as intermediate reduction states.
    For simplicity each effect declares a single operation, which eliminates
    operation tags from the syntax without affecting expressive power
    (multi-operation effects can be encoded by a sum-typed argument).

    \subsection{Operational semantics}

    Evaluation contexts come in two flavors.
    General contexts $E$ contain a hole and may cross any binder-free construct.
    \emph{Pure} contexts $P_m$, indexed by a marker $m$, additionally forbid crossing
    a $\keyword{handler}_{m'}$ delimiter for the \emph{same} marker $m$, but may cross
    delimiters for other markers as well as all administrative constructs.
    This is essential for the operation-lookup rule of multi-prompt delimited control.

    The interesting head-reduction rules are:
    \begin{align*}
        \keyword{handle}\,E\,\overline{\tau}\,t_{op}\,\keyword{in}\,t
            &\;\rightsquigarrow\; \keyword{handler}_m\,(t[\keyword{cap}_E^m\,\overline{\tau}\,t_{op}/x])
                & (m \text{ fresh}) \\
        \keyword{handler}_m\,v &\;\rightsquigarrow\; v \\
        \keyword{handler}_m\,(P_m[\keyword{perform}\,(\keyword{cap}_E^m\,\overline{\tau}\,t_{op})\,\overline{\sigma}\,v])
            &\;\rightsquigarrow\; t_{op}[\overline{\sigma}/\overline{\beta}][v,\,\keyword{resume}_m\,P_m[\cdot]] \\
        (\keyword{resume}_m\,b)\,v &\;\rightsquigarrow\; \keyword{handler}_m\,(b[v/x])
    \end{align*}
    together with the standard $\beta$, type-$\beta$, lifetime-$\beta$, and match
    rules.
    The third rule is the only one whose left-hand side is not in head position;
    it scans the pure context $P_m$ to locate the matching capability and reifies the
    captured continuation as $\keyword{resume}_m$.

    Two design points deserve emphasis.
    First, the handlers are \emph{deep}: the fourth rule re-installs the delimiter
    $\keyword{handler}_m$ around the resumed body, so a resumption invoked inside the
    operation body is itself handled by the same handler.
    This matches the discipline of Eff, Koka, and Effekt and is exactly what the state
    example of Section~\ref{sec:evaluation} relies on.
    Second, we model the ``capture up to the matching handler'' of multi-prompt
    delimited control by \emph{marker-indexed pure contexts} rather than by an explicit
    continuation stack or by named \texttt{shift}/\texttt{reset} prompts.
    The pay-off is specific to the mechanization: because markers never bind program
    variables, the reduction relation stays first-order, capture-avoiding substitution
    is the ordinary de~Bruijn operation, and the entire metatheory avoids nominal
    infrastructure (Section~\ref{sec:mechanization}).
    The resumption is kept first-class --- a value of function type --- so that the
    canonical-forms lemma for arrows acquires a second case
    (Lemma~\ref{lem:canonical}); this is the one genuinely new shape relative to a
    handler-free calculus.

    \subsection{Type system}

    The typing judgment $\Gamma \vdash t : \tau$ refines the typing of $Core_\Delta$
    with three new rules for handlers.
    Rule $T\!\_Cap$ types a runtime capability as $K_E\,local\,\overline{\tau}$.
    Rule $T\!\_HandlerM$ types $\keyword{handler}_m\,t$ at the type of $t$ and imposes
    no constraint on $m$.
    Rule $T\!\_Resume$ types $\keyword{resume}_m\,b$ as a function whose argument type
    is the resumed value's type and whose result type is the answer type recorded
    in~$b$.
    The source-level rules $T\!\_Handle$ and $T\!\_Perform$ look up the effect
    declaration in $\Gamma$ and instantiate its operation signature.

    The escape discipline of handlers is fixed by two deliberate annotation choices,
    which together implement the slogan ``$local$ may flow into a handler but never out''.
    The operation's polymorphic type arguments are bounded by $\keyword{Any}\,'\!free$,
    so an operation \emph{argument} may not itself carry a $local$ lifetime; dually,
    rule $T\!\_Resume$ types the resumption at a $local$ closure lifetime, so the
    resumption is itself confined and cannot be returned through any function boundary.
    Relaxing the first choice to admit $local$ operands is possible but requires
    propagating the operand's lifetime into the resumption and answer types; we keep
    the conservative choice here and return to the trade-off in
    Section~\ref{sec:discussion}.

    The subtyping relation is unchanged: $free <: \Delta <: local$, plus the structural
    rules for function types (contravariant on arguments, covariant on result) and
    for constructor types (covariant on lifetimes), and the $SubAny$ rule, which
    permits $\tau <: \keyword{Any}\,\Delta$ whenever every lifetime occurring in $\tau$
    outlives $\Delta$.

    \subsection{Meta-theorems and proof outlines}

    We prove the standard syntactic type-safety theorems.
    Let $\Gamma$ be an \emph{evaluation context}, that is, a typing context that
    contains neither term, type, nor effect bindings (it may still contain lifetime
    bounds and constructor declarations).

    The proofs rely on a small set of auxiliary lemmas; we state the principal ones
    explicitly here, both because they are independently useful and because their
    structure illustrates how the existential-lifetime machinery interacts with the
    handler semantics.

    \begin{lemma}[Canonical forms]\label{lem:canonical}
        Let $\Gamma$ be an evaluation context and $v$ a value with $\Gamma \vdash v : \tau$.
        \begin{enumerate}
            \item If $\tau = \overline{\sigma}\,\Delta \to \rho$, then either
                $v = \lambda(\overline{x}:\overline{\sigma'}).\,t$ for some $t$ or
                $v = \keyword{resume}_m\,b$ for some $m, b$.
            \item If $\tau = K\,\Delta\,\overline{\sigma}$, then $v =
                K'\,\Delta'\,\overline{\Delta'}\,\overline{\sigma'}\,\overline{v}$ for
                some constructor $K'$ and arguments.
            \item If $\tau = \forall l.\,\rho$, then $v = \Lambda l.\,t$.
            \item If $\tau = \forall (\alpha <: \sigma).\,\rho$, then
                $v = \Lambda(\alpha <: \sigma').\,t$.
        \end{enumerate}
    \end{lemma}

    \begin{theorem}[Progress]\label{thm:progress}
        If $\Gamma$ is an evaluation context and $\Gamma \vdash t : \tau$, then either
        $t$ is a value or there exists $t'$ such that $t \rightsquigarrow t'$.
    \end{theorem}

    \begin{proof}[Proof sketch]
        By induction on the derivation of $\Gamma \vdash t : \tau$.
        The cases for variables, $\keyword{handle}$, and $\keyword{perform}$ are
        discharged by the assumption that $\Gamma$ contains no term, type, or
        effect bindings.
        The application, type-application, lifetime-application, and \texttt{match}
        cases use Lemma~\ref{lem:canonical} together with the small-step rules of
        Figure~\ref{fig:syntax}.
        The $T\!\_HandlerM$ case branches on whether the body is a value
        (in which case $H\!\_Return$ applies) or steps internally
        (in which case the congruence $S\!\_HandlerM$ applies).
        Constructor terms require an auxiliary ``split or step'' lemma showing
        that a list of typed terms either consists entirely of values or contains
        a left-most reducible position.
    \end{proof}

    \begin{theorem}[Preservation]\label{thm:preservation}
        If $\Gamma$ is an evaluation context, $\Gamma \vdash t : \tau$, and
        $t \rightsquigarrow t'$, then $\Gamma \vdash t' : \tau$.
    \end{theorem}

    \begin{proof}[Proof sketch]
        By induction on $\Gamma \vdash t : \tau$, using the standard substitution
        lemmas for term, type, and lifetime variables.
        The non-routine cases are:
        \emph{(i)} the \texttt{match}-yes case, which discharges the
        existential-lifetime elimination via a dedicated
        $\mathit{match\_yes\_preservation}$ lemma that relates the variance-directed
        elimination function on types to the substitution induced by reduction;
        \emph{(ii)} the $H\!\_Perform$ case, which uses a lemma stating that no
        well-typed $\keyword{perform}\,(\keyword{cap}_E^m\,\ldots)$ can occur at top
        level under an evaluation context (this discharges the source-level
        $T\!\_Handle$ and $T\!\_Perform$ rules in the preservation proof).
    \end{proof}

    \begin{corollary}[Type safety]\label{cor:safety}
        If $\Gamma$ is an evaluation context, $\Gamma \vdash t : \tau$, and
        $t \rightsquigarrow^{*} t'$, then $t'$ is not stuck.
    \end{corollary}

    \begin{proof}
        By induction on the multi-step reduction, applying
        Theorems~\ref{thm:progress} and~\ref{thm:preservation} at each step.
    \end{proof}

    \subsection{Escape guarantees}\label{subsec:escape}

    Type safety is the means, not the end: the calculus exists to guarantee that
    tracked entities --- capabilities and the $local$-confined values they protect ---
    do not escape their scope. We mechanize two such guarantees.

    The first concerns runtime capabilities. Recall that the typing rule $T\!\_Cap$
    is the only rule that consults the effect environment, and that an evaluation
    context $\Gamma$ contains no effect bindings. Since every evaluation-context frame
    types its hole in the \emph{same} context (no frame introduces a binder), a
    capability sitting at an evaluation position would force its effect tag into
    $\Gamma$ --- which is impossible. This yields the strongest possible statement:
    such a configuration is not even well-typed.

    \begin{theorem}[Capability confinement]\label{thm:confine}
        Let $\Gamma$ be an evaluation context and $E$ an evaluation context frame.
        Then there is no type $\tau$ with
        $\Gamma \vdash E[\keyword{cap}_{E'}^m\,\overline{\sigma}\,t_{op}] : \tau$.
        Consequently, if $\Gamma \vdash t : \tau$ and
        $t \rightsquigarrow^{*} E[\keyword{cap}_{E'}^m\,\overline{\sigma}\,t_{op}]$,
        then this reduct is unreachable: a closed well-typed program never exposes a
        capability at an evaluation position outside its handler.
    \end{theorem}

    \begin{proof}[Proof sketch]
        The first part is a structural induction on $E$ (lemma
        $\mathit{plug\_typing\_inv}$) composed with inversion of the
        $\keyword{perform}$/$\keyword{cap}$ typing and \emph{eval\_ctx\_no\_eff}.
        The second part transports the first along preservation
        (Theorem~\ref{thm:preservation}).
    \end{proof}

    The second guarantee is the value-level counterpart of the $local \not<: free$
    lattice fact. A subsumption can never relax a $local$ lifetime to $free$; carried
    along the dynamics by preservation, this means a value produced at an escapable
    data type cannot be a $local$-confined datum in disguise.

    \begin{theorem}[Operational non-escape of $local$ values]\label{thm:nonescape}
        Let $\Gamma$ be an evaluation context and $K \neq \keyword{Any}$.
        If $\Gamma \vdash t : K\,'\!free\,\overline{\sigma}$,
        $t \rightsquigarrow^{*} v$, and $v$ is a value, then
        $v = K\,'\!\Delta\,\overline{\Delta'}\,\overline{\sigma}\,\overline{v}$ with
        $\Delta$ free of any top-level $local$.
    \end{theorem}

    \begin{proof}[Proof sketch]
        Preservation gives $\Gamma \vdash v : K\,'\!free\,\overline{\sigma}$; canonical
        forms fixes the head constructor; inversion of the constructor typing together
        with the downward-closure of the lifetime order (lemma
        $\mathit{lt\_sub\_no\_local\_mono}$) discharges the lifetime side condition.
    \end{proof}


    \section{Mechanization in Rocq/Coq} \label{sec:mechanization}

    The artifact\footnote{\url{https://github.com/penguin-boxx/effect-proofs}}
    is a Rocq/Coq development of approximately $6\,500$ lines, organized into eight
    modules (Table~\ref{tab:modules}).
    All binders use a pure de~Bruijn representation; both substitution and shifting
    are defined as ordinary recursive functions.

    \begin{table}[h]
        \centering
        \small
        \caption{Modules of the Rocq development.}
        \label{tab:modules}
        \begin{tabular}{@{}lrl@{}}
            \toprule
            \textbf{Module} & \textbf{LOC} & \textbf{Content} \\
            \midrule
            \texttt{Syntax.v}              & 133   & lifetimes, types, terms, values, decidable $\texttt{is\_value}$ \\
            \texttt{Substitution.v}        & 498   & shifting and substitution on lifetimes, types, terms \\
            \texttt{SubstitutionTheory.v}  & 972   & fusion, commutation, and zero-shift lemmas \\
            \texttt{Semantics.v}           & 602   & evaluation contexts, $\keyword{pure}_m$, head-step, full-step \\
            \texttt{Typing.v}              & 804   & lifetime subtyping, type subtyping, typing relation \\
            \texttt{Sugar.v}               & 470   & derived combinators and surface notation \\
            \texttt{Safety.v}              & 1\,863  & inversion, narrowing, canonical forms, progress, preservation,
                                                       type safety, confinement, non-escape \\
            \texttt{Examples.v}            & 1\,156  & worked examples and effect-handler programs \\
            \midrule
            \textbf{Total}                 & \textbf{6\,498} & \\
            \bottomrule
        \end{tabular}
    \end{table}

    \subsection{Representation choices}

    \paragraph{Lifetimes as a free term.}
    Lifetimes are represented as an inductive type
    $\{\,lt\_var\,n,\,lt\_free,\,lt\_local,\,lt\_min\,\}$.
    Unlike the normalized representation used in the Haskell prototype of the previous
    paper~\cite{stoyan2025existential}, the Rocq formalization keeps the syntactic
    structure of $\Delta_1 + \Delta_2$ explicit; semantic equivalence is mediated by the
    subtyping relation $\Gamma \vdash \Delta_1 <: \Delta_2$.
    This matches the on-paper presentation and avoids carrying a normalization invariant
    inside the inductive typing predicate.

    \paragraph{Capabilities reuse constructor machinery.}
    The capability constructor for an effect declaration is registered in $\Gamma$ as a
    special constructor binding $K_{cap_E}$, structurally indistinguishable from a
    user-declared data constructor.
    This dramatically reduces the number of inversion and canonical-form lemmas that
    have to be proven separately for effects, at the cost of a side condition on
    $T\!\_Ctor$ and $T\!\_Match$ ensuring that capability tags cannot be matched by
    ordinary patterns.

    \paragraph{Marker freshness.}
    Markers are natural numbers introduced by $H\!\_Handle$.
    Their freshness is not enforced syntactically; instead, the meta-theorems
    quantify universally over the marker, and the operational semantics is closed
    under $\alpha$-equivalent renaming of markers because $\keyword{handler}_m$ does
    not bind anywhere.
    This keeps the proofs first-order and avoids the need for nominal infrastructure.

    \subsection{Scope of axiomatization}

    The development closes the main meta-theorems --- progress, preservation, type
    safety, capability confinement, and operational non-escape --- on the typing and
    reduction relations, modulo a set of auxiliary lemmas posed as \texttt{Axiom}s
    rather than proved (Table~\ref{tab:axioms}).
    Two facts that earlier drafts axiomatized are now \emph{theorems}: the
    $\forall$-subtyping inversion lemma $\mathit{sub\_ty\_all\_inv\_full}$ is proved via
    a full Kernel-$F_{<:}$ narrowing development, and the
    ``no well-typed $\keyword{perform}$ of a capability survives without its handler''
    invariant is now discharged by the same structural argument that underlies
    capability confinement (Theorem~\ref{thm:confine}), rather than being assumed.
    The remaining axioms fall into two groups: standard de~Bruijn substitution
    plumbing (including its interaction with subtyping), and two domain-specific facts
    about constructor argument lists.
    The five top-level theorems use these axioms transitively.
    Two \emph{surface-syntax typing examples} in \texttt{Examples.v}
    (\texttt{typed\_match\_pair} and \texttt{typed\_choice}) are currently left
    \texttt{Admitted}; they are convenience restatements whose proof pattern is already
    exhibited by the corresponding fully-proved examples (\texttt{ex\_match\_pair} and
    the reduction lemma \texttt{red\_choice}), and no meta-theorem or reduction trace
    depends on them.

    \begin{table}[h]
        \centering
        \small
        \caption{Explicit axioms in the development, grouped by purpose.}
        \label{tab:axioms}
        \begin{tabular}{@{}p{0.32\linewidth}p{0.60\linewidth}@{}}
            \toprule
            \textbf{Group} & \textbf{Representative axioms} \\
            \midrule
            de~Bruijn substitution plumbing &
                term/type/lifetime substitution preservation
                (\texttt{subst\_tm\_lemma}, \texttt{subst\_ty\_in\_tm\_lemma},
                \texttt{subst\_lt\_in\_tm\_lemma}, and their parallel-substitution
                variants \texttt{subst\_list\_*}); shift/subst commutation
                (\texttt{shift\_subst\_lt\_*}); single-step weakening
                (\texttt{sub\_weaken\_cons}) \\[0.3em]
            Subtyping stability under substitution / shift &
                \texttt{sub\_subst\_ty}, \texttt{sub\_subst\_lt},
                \texttt{sub\_subst\_lt\_at}, \texttt{lt\_sub\_shift\_lt},
                \texttt{lt\_sub\_subst\_lt}, \texttt{shift\_lt\_sub} \\[0.3em]
            Domain-specific (constructor lists) &
                \texttt{progress\_ctor\_aux} (elementwise lifting of progress to a
                list of constructor arguments); \texttt{ctor\_step\_preservation}
                (preservation for the congruence step on constructor arguments);
                with the bridging facts \texttt{ctor\_lts\_chain\_bounded} and
                \texttt{inst\_ctor\_type\_subst\_eq} relating schema instantiation to
                iterated lifetime substitution \\
            \bottomrule
        \end{tabular}
    \end{table}

    \subsection{Build and reproducibility}

    The artifact compiles with Rocq~9.x against the Stdlib distribution.
    A top-level \texttt{Makefile} generates the Rocq make-file from \texttt{src/\_CoqProject}
    and builds all eight modules.
    On a contemporary workstation the full build takes well under one minute, which makes
    the development suitable for use as a teaching artifact or as a regression-testing
    baseline for further extensions.
    No external libraries beyond Stdlib are required.
    The trusted base is correspondingly small: the proofs use neither classical logic
    nor functional extensionality, and \texttt{Print Assumptions} on each of the five
    top-level theorems reports exactly the axioms of Table~\ref{tab:axioms} and nothing
    else --- in particular, no \texttt{Admitted} goal is reachable from any
    meta-theorem or reduction trace.


    \section{Evaluation through worked examples} \label{sec:evaluation}

    The \texttt{Examples.v} module of the development contains more than thirty
    machine-checked examples, structured into three groups.

    \paragraph{Group 1: original $Core_\Delta$ examples.}
    The eight worked examples from the original $Core_\Delta$
    paper~\cite{stoyan2025existential} are reproduced both with the primitive de~Bruijn
    syntax and with the surface-notation module from \texttt{Sugar.v}.
    These include the identity function, subsumption from $free$ to $local$ closures,
    higher-order application, lifetime polymorphism, type polymorphism, pair
    constructors at $free$ and $local$, pair destructuring, and the
    repository / make-repository / compose examples from the
    motivating section of the paper.
    All of them are checked by Rocq.

    \paragraph{Group 2: classical effect-handler programs.}
    Three handler programs --- a \emph{reader} (single \texttt{ask}), an \emph{exception}
    handler (single \texttt{throw}, the resumption discarded), and a
    \emph{non-deterministic choice} handler (a single-shot instance that resumes once
    with the chosen value) --- are both typed and reduced.
    The reduction lemmas \texttt{red\_reader}, \texttt{red\_exception}, and
    \texttt{red\_choice} witness the full reduction trace inside Rocq, and
    \texttt{typed\_reader} / \texttt{typed\_exception} establish the corresponding
    typings; the operation-polymorphic typing of the choice handler is the deferred
    example noted in Section~\ref{sec:mechanization}.

    \paragraph{Group 3: state effect with two operations.}
    A state effect with \texttt{get} and \texttt{put} is encoded via two nested handlers,
    and three reduction traces (\texttt{red\_state\_get}, \texttt{red\_state\_put},
    \texttt{red\_state\_putget}) confirm that the marker-indexed pure-context machinery
    correctly threads the state through nested handlers.
    These examples are the strongest tests of the operational semantics, as they exercise
    both nested handlers and resumptions invoked across them.

    \paragraph{Group 4: escape guarantees in action.}
    A small group of examples instantiates the guarantees of
    Section~\ref{subsec:escape} on concrete programs.
    The lemma \texttt{cap\_escape\_blocked} witnesses, at the type level, that a
    capability type $E\,'\!local\,\overline{\tau}$ fails the $no\_local$ side condition
    that $T\!\_Lam$ imposes on function results --- so a handler body can never return
    its capability or its resumption.
    Together with the operational theorems
    (\texttt{capability\_never\_exposed},
    \texttt{local\_value\_does\_not\_escape}), these checks confirm that the design
    forbids exactly the leaks it is meant to forbid, and that it does so without
    rejecting any of the thirty-odd positive programs above.

    \paragraph{Discussion.}
    The examples serve a dual purpose.
    First, they sanity-check the operational semantics: any change to
    \texttt{Semantics.v} that breaks delimited-control behavior is caught immediately
    by the reduction lemmas.
    Second, they demonstrate that the typing rules can be inhabited at the granularity
    required by realistic programs --- not just by the smallest possible test cases.
    The fact that the same artifact mechanizes both the syntactic safety theorems
    \emph{and} non-trivial reduction examples gives substantially more confidence in
    the design than either ingredient would provide alone.


    \section{Discussion} \label{sec:discussion}

    \subsection{Design decisions and trade-offs}

    The combination of pure de~Bruijn indices with marker-indexed evaluation contexts
    was chosen for two reasons.
    First, it avoids the need for nominal infrastructure (e.g.\ Aydemir et al.'s
    \texttt{Metalib}), which considerably lowers the maintenance cost of the development.
    Second, since markers never bind program variables, they do not introduce the
    classical $\alpha$-conversion difficulties that motivate nominal techniques.
    The price paid is that marker freshness is enforced semantically (by quantifying
    universally over markers in the meta-theorems) rather than syntactically, which
    is acceptable in our setting but would not scale to systems where markers
    \emph{do} interact with binders.

    Restricting effects to a single operation simplifies the syntax substantially
    without losing expressive power, as multi-operation effects can be encoded by
    a sum-typed argument and a $\keyword{match}$ inside the operation body.
    The price is some boilerplate at the user level; we judge this an acceptable
    trade-off for an artifact whose primary purpose is to deliver a sharp soundness
    proof.

    A further deliberate choice concerns the lifetime of operation arguments.
    As noted in Section~\ref{sec:formalization}, operation arguments are bounded so as
    to be $free$, while resumptions are typed at a $local$ closure lifetime.
    This realizes a clean asymmetry --- ``$local$ may flow into a handler but never
    out'' --- and is precisely what makes the confinement and non-escape theorems go
    through unconditionally.
    Admitting $local$ operands (so that a handler could retain a tracked value for the
    dynamic extent of its handler) is an attractive generalization, but it is not free:
    the operand's lifetime would have to be propagated into the resumption and answer
    types, and the confinement proof would acquire an additional case showing that an
    inward-flowing $local$ value is either consumed before the body returns or threaded
    back through a re-installed handler. We leave this extension to future work and
    flag it as the natural next increment in expressive power.

    \subsection{Limitations and threats to validity}

    Three limitations are worth making explicit.
    \begin{enumerate}
        \item \emph{Axiomatized lemmas.}
            As discussed in Section~\ref{sec:mechanization}, a small set of
            auxiliary lemmas remains axiomatized.
            Since earlier drafts, the $\forall$-subtyping inversion lemma and the
            ``no escaping $\keyword{perform}$ of a capability'' invariant have been
            fully proved (the latter is now the engine of capability confinement);
            what remains is concentrated in de~Bruijn substitution plumbing and two
            facts about constructor argument lists, both standard but voluminous to
            discharge by hand.
            Two convenience \emph{typing} examples are additionally left
            \texttt{Admitted}; no meta-theorem or reduction trace depends on them.
            The axioms are explicit and isolated; nevertheless, fully closing them
            is the most important next step.
        \item \emph{Single-operation effects.}
            The system handles only one operation per effect declaration; while
            multi-operation effects are encodable, an unconstrained extension would
            improve usability and is on the immediate roadmap.
        \item \emph{No mechanized link to the type-checker prototype.}
            The Haskell prototype of $Core_\Delta$~\cite{stoyan2025existential} and the
            Rocq specification are currently kept in sync by hand.
            Extracting a certified reference algorithm from the Rocq definitions would
            close this gap and is also planned.
    \end{enumerate}

    \subsection{Implications for language design}

    Beyond the immediate technical results, the mechanization offers two implications
    for language design.
    First, it confirms that an existential-lifetime escape analysis with an explicit
    minimum can be combined with first-class effect handlers without sacrificing
    type safety \emph{or} capability confinement, which provides a concrete foundation
    for designing future capability-based effect systems with simple user-facing
    annotations.
    Second, the explicit comparison of mechanization techniques (pure de~Bruijn vs.\
    nominal) and the precise demarcation of axiomatized lemmas can serve as a
    template for future mechanizations of similar calculi.


    \section{Conclusion} \label{sec:conclusion}

    We have presented a mechanized type-safety proof, in Rocq/Coq, for the
    $Core_\Delta$ escape-analysis calculus extended with single-operation algebraic
    effect handlers.
    The development covers syntax, small-step semantics with marker-indexed pure
    evaluation contexts, a subtyping relation over a two-point lifetime lattice with
    minimum, and a typing relation that re-binds existential lifetimes on pattern
    matching.
    Progress and preservation are proved as fully closed theorems modulo a small,
    explicitly delimited set of axiomatized auxiliary lemmas; type safety follows as
    a corollary.
    On top of safety we mechanize the two escape guarantees that motivate the calculus:
    a capability is never well-typed at an evaluation position outside its handler, and
    a value delivered at an escapable $free$ type never carries a $local$-confined
    lifetime.
    A library of more than thirty examples --- including state, exceptions, and
    non-deterministic choice --- exercises both the typing and the reduction relations.
    To the best of our knowledge, this is the first mechanized soundness proof to
    combine existential-lifetime escape analysis with first-class effect handlers.

    Future work proceeds along three axes.
    First, the remaining axiomatized lemmas --- now confined to de~Bruijn substitution
    plumbing and two constructor-list facts, the $\forall$-subtyping inversion having
    already been discharged --- need to be closed, along with the two deferred typing
    examples.
    Second, effects should be generalized to multi-operation declarations, which
    requires reinstating operation tags but is otherwise routine, and operation
    arguments should be allowed to carry $local$ lifetimes (the ``$local$ in, never
    out'' extension of Section~\ref{sec:discussion}), which sharpens the confinement
    theorem rather than merely enlarging the syntax.
    Third, the mechanized calculus should be connected with the Haskell prototype of
    the type-checker by extracting a certified reference algorithm from the Rocq
    definitions, thereby closing the gap between the formal specification and the
    executable implementation.


    \bibliographystyle{plain}
    \bibliography{bib}

    % ---------------------------------------------------------------------
    % Russian summary block, placed at the end as required by the journal.
    % ---------------------------------------------------------------------
    \selectlanguage{russian}

    \vspace{1.5em}
    \noindent\textbf{УДК} 004.052.42, 004.43

    \vspace{0.5em}
    \begin{center}
        {\large\bfseries
            Машинно-проверенное доказательство типобезопасности исчисления
            escape-анализа с экзистенциальными временами жизни и обработчиками эффектов\par}

        \vspace{0.6em}
        {\normalsize\textbf{А.\,Стоян}\textsuperscript{1}\par}

        \vspace{0.4em}
        {\small\itshape
            \textsuperscript{1}\,Национальный исследовательский университет <<Высшая школа экономики>>\\
            (НИУ ВШЭ), ул.~Кантемировская, 3А, Санкт-Петербург, 194100, Российская Федерация\par}
    \end{center}

    \vspace{0.5em}
    \noindent\textbf{Аннотация.}
    Типизированный escape-анализ~--- это техника статического анализа, поднимающая в систему
    типов вопрос о том, не переживает ли значение свою лексическую область видимости;
    она лежит в основе стековой аллокации, заимствования во владельческих языках и обеспечения
    безопасности систем эффектов на основе возможностей (capabilities).
    Недавно предложенное исчисление $Core_\Delta$ снижает нагрузку на программиста за счёт
    сочетания трёх проектных решений: \emph{экзистенциальных} времён жизни в составных типах
    данных, явной операции \emph{минимума} над временами жизни и ограниченного полиморфизма с
    границами, аннотированными временами жизни.
    Однако метатеоретические гарантии $Core_\Delta$ до сих пор обосновывались лишь на бумаге,
    а главное приложение~--- обработчики алгебраических эффектов на основе capabilities~---
    обсуждалось неформально.
    \emph{Цель работы}~--- закрыть оба пробела машинно-проверенным доказательством
    типобезопасности для исчисления, объединяющего экзистенциальные времена жизни с
    делимитированно-управляющими обработчиками эффектов.
    \emph{Методы.}
    Исчисление $Core_\Delta$ расширяется одно-операционными алгебраическими обработчиками
    эффектов в стиле передачи capability; для него задаётся малошаговая операционная семантика
    на основе контекстов вычислений, индексированных маркерами, реифицирующих захваченные
    продолжения в качестве полноправных значений времени выполнения, после чего проводится полная
    механизация~--- синтаксиса, семантики, отношения подтипизации и отношения типизации~--- в
    средствe доказательств Rocq/Coq с использованием индексов де~Брёйна.
    \emph{Результаты.}
    Доказаны классические теоремы продвижения (progress) и сохранения (preservation) для
    расширенного исчисления; в качестве следствия получена типобезопасность.
    Помимо типобезопасности, машинно доказаны две гарантии escape-анализа, ради которых
    и создавалось исчисление: теорема об \emph{удержании возможностей} (capability
    confinement), согласно которой возможность времени выполнения никогда не может
    оказаться в вычислительной позиции вне своего обработчика, и теорема об
    \emph{операционном неускользании}, согласно которой значение, выдаваемое на
    ускользающем ($free$) типе данных, не может нести ограниченное ($local$) время жизни.
    Разработка содержит около $6\,500$ строк кода на Rocq, организованных в восемь модулей,
    и сопровождается более чем тридцатью машинно-проверенными примерами, включая кодирования
    состояния, исключений и недетерминированного выбора в виде алгебраических эффектов,
    редуцируемыми непосредственно внутри ассистента доказательств.
    Сравнение со смежными работами показывает, что, насколько нам известно, представленная
    разработка является первым машинно-проверенным доказательством типобезопасности,
    действительно объединяющим оба ингредиента~--- escape-анализ на экзистенциальных временах
    жизни и обработчики эффектов первого класса~--- в рамках единого формализма.
    \emph{Практическая значимость.}
    Артефакт предоставляет переиспользуемое, машинно-проверенное основание для проектирования
    типизированных escape-анализов в языках с обработчиками эффектов первого класса; он
    открыто доступен по адресу \url{https://github.com/penguin-boxx/effect-proofs}.

    \vspace{0.4em}
    \noindent\textbf{Ключевые слова:}
    системы типов, escape-анализ, экзистенциальные времена жизни, алгебраические обработчики
    эффектов, делимитированное управление, удержание возможностей, машинная метатеория,
    Rocq/Coq, типобезопасность.

    \vspace{0.4em}
    \noindent\textbf{Благодарности.}
    Автор благодарит коллег из НИУ ВШЭ за плодотворные обсуждения проектирования систем
    эффектов на основе capabilities.
    Финансирование исследования из внешних источников не привлекалось.

\end{document}
